---
title: "1. Quick-Start: Forest Plots for NONMEM Models"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{1. Quick-Start: Forest Plots for NONMEM Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(dplyr)
  library(ggplot2)
})

theme_set(theme_bw(base_size = 14))
theme_update(plot.title = element_text(hjust = 0.5))

set.seed(865765)
```

# Introduction

This vignette is the shortest path from a fitted NONMEM model to a Forest plot. It
uses a PK model estimated with PsN and produces a parametric Forest plot of the
covariate effects on clearance, with the uncertainty taken from the NONMEM
covariance step.

A parametric Forest plot is built in five steps:

1. Describe the covariate values to display (`setupDfCovs()`).
2. Write the function that turns parameter estimates and covariate values into the
   quantities to plot.
3. Draw parameter vectors that carry the estimation uncertainty (`getSamples()`).
4. Compute the plot data (`getForestDFSCM()`).
5. Draw the plot (`forestPlot()`).

The [Walkthrough](Part2-walkthrough.html) covers the same steps in full, including
empirical Forest plots, alternative uncertainty sources, an explicit reference
subject, and the formatting arguments of `forestPlot()`.

# 1. Describe the covariates

`setupDfCovs()` reads an analysis data set, summarises each covariate, and returns
a data frame in which every row is one row of the Forest plot. Continuous
covariates are represented by their 5th and 95th percentiles; binary covariates
are expanded to one row per level.

```{r}
dataFile <- system.file("extdata", "SimVal/DAT-1-MI-PMX-2.csv", package = "PMXForest")
dfData   <- read.csv(dataFile)

covariates <- c("WT", "AGE", "CRCL", "SEX", "NCIL", "FORM", "FOOD", "RACEL2")

dfCovs <- setupDfCovs(data = dfData, covariates = covariates, idVar = "ID")
dfCovs
```

Cells that are inactive on a row hold the missing-value indicator `-99`; the
parameter function in the next step treats that as "no effect". Multi-level
categorical covariates (one-hot encoding, reference level) are covered in the
[Walkthrough](Part2-walkthrough.html).

# 2. Write the parameter function

The function takes a vector of `thetas` (in the order the THETAs appear in the
NONMEM model) and a one-row data frame `df` of covariate values. It rebuilds the
relevant part of the model and returns the quantities to plot.

Every covariate test guards for the missing-value indicator (`-99`), which
`setupDfCovs()` places in the cells that are inactive on a given row, and for the
covariate column being absent entirely.

```{r}
paramFunction <- function(thetas, df, ...) {

  CLFOOD <- 1
  if (any(names(df) == "FOOD") && df$FOOD != -99 && df$FOOD == 0) CLFOOD <- 1 + thetas[11]

  FRELFORM <- 1
  if (any(names(df) == "FORM") && df$FORM != -99 && df$FORM == 0) FRELFORM <- 1 + thetas[12]

  FRELSEX <- 1
  if (any(names(df) == "SEX") && df$SEX != -99 && df$SEX == 2) FRELSEX <- 1 + thetas[14]

  FREL <- thetas[1] * FRELSEX * FRELFORM

  if (any(names(df) == "WT") && df$WT != -99) {
    TVCL <- thetas[4] * (df$WT / 75)^thetas[2]
  } else {
    TVCL <- thetas[4]
  }
  CL <- CLFOOD * TVCL

  if (any(names(df) == "WT") && df$WT != -99) {
    V <- thetas[5] * (df$WT / 75)^thetas[3]
  } else {
    V <- thetas[5]
  }

  AUC <- 80 / (CL / FREL)

  list(CL, FREL, AUC, V)
}

functionListName <- c("CL", "Frel", "AUC", "V")
```

# 3. Sample the parameter uncertainty

`getSamples()` reads the `.ext` file for the final estimates and the `.cov` file
for the covariance matrix, then draws `n` parameter vectors from the implied
multivariate normal distribution.

```{r}
extFile <- system.file("extdata", "SimVal/run7.ext", package = "PMXForest")
covFile <- system.file("extdata", "SimVal/run7.cov", package = "PMXForest")

dfSamples <- getSamples(covFile, extFile = extFile, n = 200)
```

# 4. Compute the Forest plot data

`getForestDFSCM()` evaluates `paramFunction` for every row of `dfCovs` and every
parameter vector in `dfSamples`, then summarises the results into the columns
`forestPlot()` needs. `noBaseThetas` is the number of THETAs in the model.

With `cdfCovsNames` left at its default, each row is labelled from its own
covariate values (`WT=58.5`, `SEX=2`, and so on). The
[Walkthrough](Part2-walkthrough.html) shows how to supply publication-quality row
labels and covariate group names.

```{r}
dfres <- getForestDFSCM(
  dfCovs           = dfCovs,
  functionList     = list(paramFunction),
  functionListName = functionListName,
  noBaseThetas     = 14,
  dfParameters     = dfSamples
)
```

# 5. Draw the plot

```{r}
forestPlot(dfres, parameters = "CL")
```

The dotted line marks the reference: here no `dfRefRow` was supplied, so the
reference is the model with every covariate effect switched off. The points show
the covariate effect on `CL` relative to that reference, and the bars its 90%
confidence interval. `WT` is fixed in this model, so its interval has zero width.

::: {.callout-note}
**Next steps.** The [Walkthrough](Part2-walkthrough.html) adds an explicit
reference subject, empirical Forest plots, bootstrap and SIR uncertainty, and the
formatting arguments of `forestPlot()`. The deep-dive vignettes cover
[preparing Forest plot inputs](Part3-deep-dive-forest-plot-inputs.html),
[R-coded models](Part3-deep-dive-r-coded-models.html), and
[time-to-event models](Part3-deep-dive-tte-models.html).
:::
