---
title: "3. Deep Dive: Forest Plots for Time-to-Event Models"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Forest Plots for Time-to-Event Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
.callout-warning {
  background-color: #fff3cd;
  border-left: 4px solid #ffc107;
  color: #664d03;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(ggplot2)
})

theme_set(theme_bw(base_size = 14))
theme_update(plot.title = element_text(hjust = 0.5))

set.seed(865765)
```

# Introduction

This vignette builds a Forest plot from a time-to-event (TTE) model. The workflow
is the parametric SCM workflow of the [Walkthrough](Part2-walkthrough.html), with
two differences specific to TTE models:

* the quantities plotted are the **hazard ratio** relative to a reference subject
  and the **probability of an event within one year**;
* computing the event probability requires translating the NONMEM
  parameterisation of the event-time distribution (Weibull here) into the
  parameterisation R uses.

The covariate effects are assumed to enter the model as proportional hazards on
the base hazard, $h_0(t) \cdot e^{(\beta_1 x_1 + \beta_2 x_2 + \dots)}$.

# 1. Covariate values

```{r}
dfCovs <- createInputForestData(
  list(
    "AGE" = c(35, 50, 65),
    "EXP" = c(0, 100, 200, 300, 400, 500)
  ),
  iMiss = -99
)
dfCovs

covnames <- c(
  "Age 35 y", "Age 50 y", "Age 65 y",
  "Placebo",
  "Drug conc 100 ng/mL", "Drug conc 200 ng/mL", "Drug conc 300 ng/mL",
  "Drug conc 400 ng/mL", "Drug conc 500 ng/mL"
)
stopifnot(length(covnames) == nrow(dfCovs))
```

The reference subject is 50 years old and in the placebo group (exposure 0).

```{r}
dfRefRow <- data.frame("AGE" = 50, "EXP" = 0)
```

# 2. Parameter functions

Two functions are used so the hazard ratio and the event probability can be given
different x-axis scales. `paramFunctionCDF()` calls `paramFunctionHR()`, and the
extra arguments it needs (`functHR`, `TIME`) are passed through from
`getForestDFSCM()`.

Both functions guard for absent covariate columns and for the `-99` indicator.
That contract — and why the `-99` guard is needed even when the data has no
missing values — is set out in
[Preparing Forest plot inputs, Section 2.1](Part3-deep-dive-forest-plot-inputs.html).

```{r}
## Hazard ratio relative to the reference subject
paramFunctionHR <- function(thetas, df, ...) {
  RF <- 0
  if (any(names(df) == "AGE") && df$AGE != -99) {
    RF <- RF + (df$AGE - 50) * thetas[3]
  }

  EFF <- 0
  if (any(names(df) == "EXP") && df$EXP != -99) {
    EFF <- thetas[4] * df$EXP
  }

  exp(RF + EFF)
}

## Probability of an event up to time TIME
paramFunctionCDF <- function(thetas, df, functHR, TIME, ...) {
  ## Exponentiated covariate and exposure effects
  COV <- functHR(thetas, df, ...)

  ## Translate the NONMEM parameterisation into the pweibull() parameterisation.
  ## This step is specific to how the model is written.
  TH1 <- 1 / thetas[1]
  TH2 <- thetas[2]
  LAM <- ((1 / COV)^(1 / TH2)) * TH1

  pweibull(TIME, TH2, LAM)
}

functionListNameHR <- c("Hazard Ratio")
functionListNameCDF <- c("Probability of an event within 1 year")
```

::: {.callout-warning}
Re-implementing the model in R duplicates code that already exists in the NONMEM
control stream. Check the translation carefully — especially the distribution
reparameterisation, which does not affect the hazard ratio but drives the event
probability.
:::

# 3. Parameter uncertainty

```{r}
runno <- "tte_weibull"
extFile <- system.file("extdata", "tte", paste0(runno, ".ext"), package = "PMXForest")
```

## From the covariance step

The model has no random effects, so the `OMEGA` and `SIGMA` samples are set to
zero.

```{r}
covFile <- system.file("extdata", "tte", paste0(runno, ".cov"), package = "PMXForest")
dfSamplesCOV <- getSamples(covFile, extFile = extFile, n = 200)
```

## From a bootstrap

```{r}
bootFile <- system.file(
  "extdata", "tte", "bootstrap_tte_weibull_n500",
  paste0("raw_results_", runno, ".csv"),
  package = "PMXForest"
)
dfSamplesBS <- getSamples(bootFile, extFile = extFile)
```

# 4. Forest plot data

`noBaseThetas` is 4 for this model. `paramFunctionCDF()` receives `functHR` and
`TIME` as extra named arguments.

```{r}
dfresHR <- getForestDFSCM(
  dfCovs           = dfCovs,
  cdfCovsNames     = covnames,
  functionList     = list(paramFunctionHR),
  functionListName = functionListNameHR,
  noBaseThetas     = 4,
  dfParameters     = dfSamplesBS,
  dfRefRow         = dfRefRow,
  cstrPackages     = c("dplyr")
)

dfresCDF <- getForestDFSCM(
  dfCovs           = dfCovs,
  cdfCovsNames     = covnames,
  functionList     = list(paramFunctionCDF),
  functionListName = functionListNameCDF,
  noBaseThetas     = 4,
  dfParameters     = dfSamplesBS,
  dfRefRow         = dfRefRow,
  cstrPackages     = c("dplyr", "stats"),
  functHR          = paramFunctionHR,
  TIME             = 365
)
```

# 5. Plots

`forestPlot(return = "plotList")` gives the panel list, so a `ggplot` scale can be
added before the panels are recombined with `ggpubr::ggarrange()`.

## Hazard ratios

The x-axis is on a `log10` scale, so a halving (HR = 0.5) and a doubling (HR = 2)
of the hazard sit the same distance from the reference.

```{r ForestHR, fig.width=10, fig.asp=0.5, out.width="100%"}
plotList <- forestPlot(
  dfresHR,
  plotRelative   = FALSE,
  xlb            = "Hazard Ratio",
  return         = "plotList",
  sigdigits      = 3,
  strip_top_size = 12
)

plotList[[1]] <- plotList[[1]] + scale_x_continuous(trans = "log10")

ggpubr::ggarrange(
  plotlist = plotList,
  ncol = 2, nrow = 1,
  widths = c(2, 0.7),
  align = "h",
  common.legend = TRUE
)
```

## Probability of an event within a year

The x-axis is shown as a percentage.

```{r ForestCDF, fig.width=10, fig.asp=0.5, out.width="100%"}
plotList <- forestPlot(
  dfresCDF,
  plotRelative     = FALSE,
  xlb              = "Probability of event",
  return           = "plotList",
  sigdigits        = 3,
  strip_top_size   = 12,
  statisticsLabel  = "Statistics"
)

plotList[[1]] <- plotList[[1]] + scale_x_continuous(labels = scales::percent)

ggpubr::ggarrange(
  plotlist = plotList,
  ncol = 2, nrow = 1,
  widths = c(2, 0.7),
  align = "h",
  common.legend = TRUE
)
```
