---
title: "3. Deep Dive: Secondary Parameters"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Secondary Parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
.callout-warning {
  background-color: #fff3cd;
  border-left: 4px solid #ffc107;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(ggplot2)
})
theme_set(theme_bw(base_size = 14))
set.seed(865765)
```

# Introduction

A Forest plot usually shows model parameters -- `CL`, `V`, a bioavailability
term. Often the quantity of clinical interest is a **secondary** parameter
derived from those: an exposure `AUC` over a dosing interval, a terminal
half-life, a steady-state `Cmax`. These are not in `$PK`; they are computed from
the structural parameters, sometimes in closed form, sometimes by solving the
model for a particular dosing design.

`createParamFunction()` translates `$PK` into an R parameter function
automatically. Its `secondary` argument attaches the derived quantities in the
same call, so the whole parameter function -- structural part and secondary
part -- is generated as one object that `getForestDFSCM()` can consume directly.

Each `secondary` entry names a variable to add to the return list; its value is
**one of**

* a line of R code -- `secondary = list(AUC = "df$DOSE / CL")`;
* the path to an `.R` file of arbitrary code, including a `deSolve` or
  `mrgsolve` simulation -- `secondary = list(CMAX = "cmax.R")`; or
* a **list** with that `source` plus any constants the code needs --
  `secondary = list(CMAX = list(source = "cmax.R", dose = 100, tau = 12))`.
  Each constant is bound (`dose <- 100`, `tau <- 12`) immediately before the
  code, inside the same `local()` block, so a reusable file is parametrised at
  the call site instead of hard-coding the regimen. A bare string is just
  `source` with no constants.

A file's text is *inlined* into the generated source, so the result stays a
self-contained artifact: it does not depend on the file still being on disk when
the parameter function later runs on a worker process.

The two parts below are independent:

* **Part 1** derives `AUC`, an elimination rate and a half-life in closed form.
  It has no compiled dependencies and runs anywhere -- the code and output on
  this page are from a live run.
* **Part 2** computes a steady-state `Cmax` with an `mrgsolve` simulation. Its
  chunks are **not executed** when this vignette is built; the outputs shown
  there are illustrative. The code is runnable if you have a working `mrgsolve`.

---

# Part 1 -- closed-form secondary parameters

## 1.1 The structural parameter function

Start from the bundled `run7` control stream and ask for the structural
parameters.

```{r}
modFile <- system.file("extdata", "SimVal/run7.mod", package = "PMXForest")
extFile <- system.file("extdata", "SimVal/run7.ext", package = "PMXForest")

base <- suppressWarnings(
  createParamFunction(modFile, parameters = c("CL", "V"), quiet = TRUE)
)
base$functionListName
```

`run7`'s `$PK` also computes `FREL` (relative bioavailability); the generator
keeps that intermediate assignment, so a secondary block spliced in afterwards
can use `FREL` by name.

## 1.2 Attach the secondary parameters

Three entries, showing each form:

* **`AUC`** -- a config list: `source` is a one-liner, and `dose` is a constant
  supplied at the call site (100 mg here). Steady-state `AUC` over a dosing
  interval for a linear model is `F * dose / CL`.
* **`KEL`** -- a bare snippet, `CL / V`.
* **`THALF`** -- another snippet that **uses `KEL`**: entries are emitted in
  order, so a later one may refer to an earlier one.

```{r}
out <- suppressWarnings(createParamFunction(
  modFile,
  parameters = c("CL", "V"),
  secondary  = list(
    AUC   = list(source = "dose * FREL / CL", dose = 100),
    KEL   = "CL / V",
    THALF = "log(2) / KEL"
  ),
  quiet = TRUE
))

out$functionListName   # CL, V, then the three secondaries
out$primaryNames
out$secondaryNames
```

The tail of the generated source -- the constant binding, then each entry
inside its own `local()` block, then the extended return list:

```{r, echo = FALSE, comment = ""}
tail(out$code, 22)
```

## 1.3 Evaluate at the final estimates

Read the function in and call it for a reference subject.

```{r}
paramFunction <- eval(parse(text = out$code))

thetas <- as.numeric(
  getExt(extFile)[getExt(extFile)$ITERATION == -1000000000, 2:15]
)

paramFunction(thetas = thetas, df = data.frame(WT = 75, FOOD = 1))
```

## 1.4 The Forest plot

`functionListName` already carries the secondary names, so `getForestDFSCM()`
returns rows for `AUC`, `KEL` and `THALF` with no extra work.

```{r}
dfData <- read.csv(
  system.file("extdata", "SimVal/DAT-1-MI-PMX-2.csv", package = "PMXForest")
)
dfCovs <- setupDfCovs(dfData, covariates = c("WT", "FOOD"), idVar = "ID")

dfSamples <- getSamples(
  system.file("extdata", "SimVal/run7.cov", package = "PMXForest"),
  extFile, n = 50
)

dfres <- getForestDFSCM(
  dfCovs           = dfCovs,
  functionList     = list(paramFunction),
  functionListName = out$functionListName,
  noBaseThetas     = out$noBaseThetas,
  dfParameters     = dfSamples
)

unique(dfres$PARAMETER)
```

```{r, fig.width = 9, fig.asp = 0.5, out.width = "100%"}
forestPlot(
  dfres[dfres$PARAMETER == "AUC", ],
  plotRelative = TRUE,
  keepYlabs    = TRUE,
  rightStrip   = FALSE
) +
  ggtitle("Steady-state AUC vs. reference")
```

---

# Part 2 -- a simulation-based secondary with mrgsolve

::: {.callout-warning}
**The chunks in this part are not executed when the vignette is built.** The
outputs shown are illustrative, not the result of a live run. The code itself
*is* runnable, but you must have a working `mrgsolve` installation first -- which
means a C++ toolchain (Rtools on Windows, the Xcode command-line tools on macOS,
`r-base-dev` / `build-essential` on Linux), because `mrgsolve` compiles each
model on first use. To run this part yourself, copy the chunks into a script and
execute them, or set `eval = TRUE` per chunk.
:::

::: {.callout-note}
**mrgsolve version.** The shipped file below uses only long-stable core API --
`mcode_cache()`, `param()`, `ev()`, `mrgsim_df()` -- and the basic
`$PARAM` / `$CMT` / `$ODE` / `$TABLE` / `$CAPTURE` blocks. It needs **no
version-specific features**. It has been checked on **mrgsolve 1.0.9**; any
**mrgsolve >= 1.0.0** should be enough. The current CRAN release is **2.0.1**
and these calls are unchanged across the 1.x -> 2.x boundary, so the latest
version is *not* required.
:::

## 2.1 The shipped secondary file

`PMXForest` ships an example at
`system.file("secondary-cmax-mrgsolve.R", package = "PMXForest")`. It compiles a
one-compartment oral model, simulates a course of oral doses, and returns the
maximum concentration in the final (steady-state) dosing interval -- the value
of its last expression becomes `CMAX`. The regimen constants `dose`, `tau` and
`n` come from the `secondary` config list (2.2); the
`exists(name, inherits = FALSE)` guards let the file fall back to defaults when a
bare-string entry supplies no constants.

```{r, echo = FALSE, comment = ""}
cmaxFile <- system.file("secondary-cmax-mrgsolve.R", package = "PMXForest")
cat(readLines(cmaxFile), sep = "\n")
```

Points worth noting:

* **Scope.** Inside the generated function the block sees `thetas`, `df`, `...`,
  the call-site constants, and every structural parameter (`CL`, `V`, `FREL`,
  `KA`, ...) by name. Covariate columns are reached as `df$NAME` -- there is no
  bare `WT`.
* **Compilation cost.** `mrgsolve::mcode_cache()` keeps the compiled model on
  disk between calls. The parameter function is evaluated once per covariate row
  per bootstrap sample -- hundreds to thousands of times -- so recompiling on
  every call would dominate the run. Cache the model, or build it once outside
  and pass it in through `...`.
* **Return a scalar.** `getForestDFSCM()` expects each returned element to be a
  single number; the last line reduces the simulation with `max(...)`.

## 2.2 Generate the combined parameter function

The config-list form passes the regimen (80 mg every 24 h, 7 doses to steady
state) into the shipped file:

```{r, eval = FALSE}
cmaxFile <- system.file("secondary-cmax-mrgsolve.R", package = "PMXForest")

out <- suppressWarnings(createParamFunction(
  modFile,
  parameters = c("CL", "V", "MAT"),
  secondary  = list(
    CMAX = list(source = cmaxFile, dose = 80, tau = 24, n = 7)
  ),
  quiet = TRUE
))

out$functionListName
#> [1] "CL"   "V"    "MAT"  "CMAX"
```

The tail of the generated source carries the constants, then the file's text
spliced in **verbatim** below them (re-indenting it would corrupt the multi-line
`mrgsolve` model string):

```r
  ## ---- Secondary parameters -----------------------------------------------
  ## CMAX  (inlined from secondary-cmax-mrgsolve.R)
  CMAX <- local({
    dose <- 80
    tau <- 24
    n <- 7
# ... file header ...
dose <- if (exists("dose", inherits = FALSE)) dose else 80
tau  <- if (exists("tau",  inherits = FALSE)) tau  else 24
n    <- if (exists("n",    inherits = FALSE)) n    else 7
.code <- "
$PARAM CL = 1, V = 10, KA = 1
...
"
.mod <- mrgsolve::mcode_cache("pmxforest_secondary_1cmt_oral", .code)
# ...
.lastInterval <- subset(.sim, time >= (n - 1) * tau)
max(.lastInterval$CP)
  })

  list(CL = CL, V = V, MAT = MAT, CMAX = CMAX)
```

Because the constants are bound *first*, the file's `exists(..., inherits =
FALSE)` guards keep the passed-in values. A bare
`secondary = list(CMAX = cmaxFile)` emits no constants and the file's defaults
(the same 80 / 24 / 7) apply.

## 2.3 Sanity-check at the final estimates

```{r, eval = FALSE}
paramFunction <- eval(parse(text = out$code))

thetas <- as.numeric(
  getExt(extFile)[getExt(extFile)$ITERATION == -1000000000, 2:15]
)

paramFunction(thetas = thetas, df = data.frame(WT = 75, FOOD = 1))
#> $CL
#> [1] 11.9074
#>
#> $V
#> [1] 191.231
#>
#> $MAT
#> [1] 6.46559
#>
#> $CMAX
#> [1] 0.3699538          # ng/mL, steady-state peak for the reference subject
```

## 2.4 Feed it to `getForestDFSCM()`

Nothing special is needed beyond loading `mrgsolve` on the worker processes,
hence `cstrPackages`.

```{r, eval = FALSE}
dfData <- read.csv(
  system.file("extdata", "SimVal/DAT-1-MI-PMX-2.csv", package = "PMXForest")
)
dfCovs <- setupDfCovs(
  dfData, covariates = c("WT", "FOOD", "FORM"), idVar = "ID"
)
dfSamples <- getSamples(
  system.file("extdata", "SimVal/run7.cov", package = "PMXForest"),
  extFile, n = 200
)

dfres <- getForestDFSCM(
  dfCovs           = dfCovs,
  functionList     = list(paramFunction),
  functionListName = out$functionListName,   # CL, V, MAT, CMAX
  noBaseThetas     = out$noBaseThetas,
  dfParameters     = dfSamples,
  cstrPackages     = "mrgsolve",
  ncores           = 1
)

unique(dfres$PARAMETER)
#> [1] CL   V    MAT  CMAX
#> Levels: CL V MAT CMAX
```

## 2.5 The Forest plot

```{r, eval = FALSE}
forestPlot(
  dfres[dfres$PARAMETER == "CMAX", ],
  plotRelative = TRUE,
  keepYlabs    = TRUE,
  rightStrip   = FALSE
) +
  ggtitle("Steady-state Cmax vs. reference")
```

---

# Notes

* **Verification.** `verifyParamFunction()` checks a generated function against a
  NONMEM `$TABLE`. It checks only the `$PK` parameters by default -- a secondary
  such as `AUC` or `CMAX` is generally not a table column and not reducible to a
  typical value -- so a function with secondaries still verifies cleanly. Pass
  `parameters = "AUC"` to force a check if you do have it tabled.
* **Snippet vs. file vs. config list.** A string is treated as a file when it
  exists on disk; a string that looks like a path (`.R` / `.r`) but does not
  exist is an error, not a silently mistyped snippet. A list entry needs a
  single-string `source`; every other element must be a named, atomic constant.
* **Constant names.** A call-site constant shadows a `$PK` variable of the same
  name inside that entry's `local()` block. Reusable files use dotted names
  (`.dose`) or the `exists(..., inherits = FALSE)` guard shown above to stay
  clear of the structural namespace.
