---
title: "2. Walkthrough: End-to-End Forest Plots for NONMEM Models"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{2. Walkthrough: End-to-End Forest Plots for NONMEM Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
.callout-warning {
  background-color: #fff3cd;
  border-left: 4px solid #ffc107;
  color: #664d03;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 9
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(dplyr)
  library(ggplot2)
})

theme_set(theme_bw(base_size = 14))
theme_update(plot.title = element_text(hjust = 0.5))

set.seed(865765)
```

# Introduction

This walkthrough covers the full process of building Forest plots from a NONMEM
model and PsN output. It extends the [Quick-Start](Part1-quick-start.html) with:

* two ways to build the covariate structure, and how to label the rows;
* the parameter function in detail, including missing-value handling;
* the four sources of parameter uncertainty `getSamples()` accepts;
* an explicit reference subject with `setupDfRefRow()`;
* *empirical* Forest plots with `getForestDFemp()`;
* the formatting arguments of `forestPlot()`.

The example is a PK model (`run7`) built with the stepwise covariate model (SCM)
in PsN. Because SCM keeps only covariates that meet a significance criterion, some
rows of a parametric Forest plot from an SCM model show no effect. An empirical
Forest plot instead reads the covariate effect from the analysis data and can
therefore show covariates that are not in the model.

```{r}
modDir <- "SimVal"
dataFile <- system.file("extdata", paste0(modDir, "/DAT-1-MI-PMX-2.csv"), package = "PMXForest")
modFile <- system.file("extdata", paste0(modDir, "/run7.mod"), package = "PMXForest")
extFile <- system.file("extdata", paste0(modDir, "/run7.ext"), package = "PMXForest")
covFile <- system.file("extdata", paste0(modDir, "/run7.cov"), package = "PMXForest")
bsFile <- system.file("extdata", paste0(modDir, "/bs7.dir/raw_results_run7bs.csv"), package = "PMXForest")
sirFile <- system.file("extdata", paste0(modDir, "/sir7.dir/raw_results_run7.csv"), package = "PMXForest")

dfData <- read.csv(dataFile)
```

# 1. Build the covariate structure

Each row of `dfCovs` becomes one row of the Forest plot. The covariate effect on
that row is a function of the covariate values in that row. Cells that are not
active on a row hold the missing-value indicator (`-99` by default).

`setupDfCovs()` builds that structure from the data. It summarises each covariate
and reshapes the result in one call, so the values plotted are the ones the study
actually observed rather than numbers typed in by hand. Continuous covariates are
represented by their 5th and 95th percentiles; categorical covariates get one row
per level.

```{r}
dfCovs <- setupDfCovs(
  data       = dfData,
  covariates = c("WT", "AGE", "CRCL", "SEX", "NCIL", "FORM", "FOOD", "GENO"),
  catRef     = list(GENO = 2),
  sep        = "",
  idVar      = "ID"
)
dfCovs
```

Two arguments are doing specific work here. The `run7` model uses genotype level
2 (2D6 EM) as its reference and estimates effects for levels 1, 3 and 4, so
`catRef = list(GENO = 2)` makes level 2 the reference — otherwise `setupDfCovs()`
drops the *lowest* level. And `sep = ""` names the one-hot columns `GENO1`,
`GENO3`, `GENO4` rather than the default `GENO_1`, matching the names the control
stream and the analysis data set both use.

::: {.callout-note}
Pass the same `catRef` and `sep` to `setupDfRefRow()` in Section 4. They have to
agree, or the reference row's columns will not line up with `dfCovs` and the
reference values are silently dropped.
:::

Row labels and covariate-group labels are supplied separately. Deriving the
numeric ones from `dfCovs` keeps them honest: `setupDfCovs()` computes the
percentiles, so hard-coding them in the labels is a standing invitation for the
two to drift apart. The `stopifnot()` makes a miscount fail immediately rather
than shifting every label below it.

```{r}
covnames <- c(
  paste0("WT ", dfCovs$WT[1:2], " kg"),
  paste0("Age ", dfCovs$AGE[3:4], " y"),
  paste0("CRCL ", dfCovs$CRCL[5:6], " mL/min"),
  "Male", "Female",
  "NCI = 0", "NCI > 0",
  "Oral tablet", "FDC",
  "Fasted", "Fed",
  "2D6 UM", "2D6 EM", "2D6 IM", "2D6 PM"
)
stopifnot(length(covnames) == nrow(dfCovs))

covariateGroupNames <- c(
  "Weight", "Age", "Creatinine\nclearance", "Sex",
  "NCI", "Formulation", "Food status", "2D6 genotype"
)
```

`setupDfCovs()` is a composition of two lower-level functions: `getCovStats()`
chooses the values, `createInputForestData()` lays them out as rows. Call them
yourself when you want values the data does not supply — a weight of exactly 70
kg, a covariate combination no subject has — or a geometry `setupDfCovs()` does
not produce. Both, along with `additionalCovs`, `useMissVal` and `contRef`, are
covered in the
[Forest plot inputs deep dive](Part3-deep-dive-forest-plot-inputs.html).

# 2. Generate the parameter function

The parameter function turns a vector of `thetas` and one row of covariate values
into the quantities to plot. Written by hand it duplicates algebra that already
exists in the control stream, and it is the easiest place in this workflow to
introduce an error that no test will catch.

`createParamFunction()` writes it by translating the model's `$PK` block.

```{r}
gen <- createParamFunction(
  modFile,
  parameters = c("CL", "FREL", "V"),
  covRef     = list(GENO1 = 0, GENO3 = 0),
  secondary  = list(AUC = "80 / (CL / FREL)"),
  quiet      = TRUE
)
```

Two arguments need explaining.

`covRef` supplies reference values the control stream does not settle on its own.
`run7`'s `$PK` carries `GENO1` and `GENO3` dummies that no `IF()` tests directly,
so the generator can only *propose* a level — and warns when it does. Stating
them removes the guess. Which covariates need this is not something to memorise:
run the call without `covRef` once and the warning names them.

`secondary` adds quantities that are not in `$PK` at all. `AUC` here is exposure
after an 80 mg dose, and the dose appears nowhere in the control stream, so no
translator could infer it. Anything derived — AUC, half-life, `Cmax` — arrives
this way; the [Secondary parameters
vignette](Part3-deep-dive-secondary-parameters.html) covers the fuller forms,
including running an `mrgsolve` or `deSolve` simulation.

**Nothing has been evaluated yet.** `gen$code` is source text, and the point of
returning text is that you read it against the control stream before trusting it.
Note the preamble: every missing-value decision is hoisted to the top and
annotated with where the reference came from, so the algebra below it is a plain
transliteration you can compare line by line with `$PK`.

```{r}
cat(gen$code, sep = "\n")
```

Once that matches the model, evaluate it:

```{r}
paramFunction <- eval(parse(text = paste(gen$code, collapse = "\n")))
```

The call also returns the two values that otherwise get typed in by hand and go
stale — the parameter names, and the THETA count `getForestDFSCM()` needs:

```{r}
functionListName <- gen$functionListName
functionListName
gen$noBaseThetas
```

If the model was run with a `$TABLE` containing both the parameters and the
covariates, `verifyParamFunction()` turns "do I trust this translation?" into a
pass or a fail, by comparing the generated function against the values NONMEM
itself wrote:

```r
verifyParamFunction(gen, tabFile = "xptab7", thetas = thetas)
#> PASS - verifyParamFunction: 3/3 parameter(s)
```

No `$TABLE` is bundled with the package, so that call is shown rather than run.

What the generated function must satisfy — the `function(thetas, df, ...)`
contract, why `df` is one row, and why the `-99` guards are needed even when the
data has no missing values — is set out in
[Preparing Forest plot inputs, Section 2.1](Part3-deep-dive-forest-plot-inputs.html).
Read it when you need to write one by hand, which Section 2.3 there describes.

# 3. Sample the parameter uncertainty

`getSamples()` returns a data frame of parameter vectors in `.ext` column order.
The confidence interval on each Forest plot row is the quantile of the parameter
function evaluated over these vectors.

```{r}
## Covariance matrix: n draws from the multivariate normal implied by the .cov step
dfSamplesCOV <- getSamples(covFile, extFile = extFile, n = 200)

## Bootstrap: the non-zero-OFV resamples from a PsN bootstrap raw_results file
dfSamplesBS <- getSamples(bsFile, extFile = extFile)

## SIR: all resamples in a PsN SIR raw_results file
dfSamplesSIR <- getSamples(sirFile, extFile = extFile)

## Small bootstrap: build a covariance matrix from the bootstrap resamples,
## then draw n vectors from it
dfSamplesBSsmall <- getSamples(bsFile, extFile = extFile, n = 200)
```

Sampling from the covariance matrix or from a small bootstrap produces symmetric
intervals. A full bootstrap or SIR file lets the interval be asymmetric.

# 4. Define the reference subject

Without a reference, `getForestDFSCM()` uses the model with every covariate effect
switched off (all cells `-99`). To make the reference a typical subject instead,
pass `dfRefRow`.

`setupDfRefRow()` builds that row, matched to the geometry of a `setupDfCovs()`
result. By default it takes the reference from the **data** — the median for a
continuous covariate, the most common level for a categorical one.

That default is the wrong choice here, and it is worth seeing why, because the
symptom is subtle. A generated parameter function falls back to the **model's**
reference when a covariate is inactive on a row: `-99` for `WT` means the
function uses 75, the normalisation constant in `$PK`. If the reference row
instead carries the data median, the two disagree, and every row where that
covariate is inactive is displaced from 1 — not because the covariate has an
effect, but because the denominator was computed differently from the numerator.

Taking both from the control stream removes the mismatch:

```{r}
refCovs <- c("WT", "AGE", "CRCL", "SEX", "NCIL", "FORM", "FOOD", "GENO")

dfRefRow <- setupDfRefRow(
  dfCovs = dfCovs,
  data = dfData,
  covariates = refCovs,
  contRef = list(WT = "model", default = "median"),
  catRef = list(
    SEX = "model", FORM = "model", FOOD = "model",
    GENO = 2, default = "mode"
  ),
  model = modFile,
  sep = "",
  idVar = "ID"
)
dfRefRow
```

Three things about that call are worth stating, because each is a trap:

**`catRef` and `sep` are repeated from Section 1 and are not optional.**
`setupDfRefRow()` keys its result by the column names `setupDfCovs()` produced,
so a different `sep` gives `GENO_1` where `dfCovs` has `GENO1`, and the reference
value for that covariate is quietly lost.

**`"model"` is given per covariate, not as a bare scalar.** As a scalar it
applies to everything, and the first covariate `$PK` does not mention — `AGE`
here — fails the whole call. The named-list form asks for `"model"` where the
model has an answer and falls back to `"median"` / `"mode"` where it does not.

**`GENO` gets a literal level, not `"model"`.** `$PK` names the dummies `GENO1`,
`GENO3`, `GENO4` and never bare `GENO`, so there is no model reference to find
for the covariate they encode. The error says so if you try it; the
[deep dive](Part3-deep-dive-forest-plot-inputs.html) shows the exact text.

A hand-built `dfCovs` needs a hand-built reference to match it — a one-row data
frame with the same columns works just as well. That route, and how to keep the
two consistent, is in the
[Forest plot inputs deep dive](Part3-deep-dive-forest-plot-inputs.html).

::: {.callout-note}
The reference enters the relative scale as the denominator: each row is plotted as
its parameter value divided by the reference-subject value. A reference subject
that carries a large covariate effect shifts every other row. Choose it
deliberately, and state it in the plot caption.
:::

# 5. Compute the parametric Forest plot data

`getForestDFSCM()` evaluates `paramFunction` for every row of `dfCovs` and every
parameter vector, then appends the summary columns `forestPlot()` reads.
`noBaseThetas` is the number of THETAs in the model. `dfCovs` here already carries
the `GENO1`/`GENO3`/`GENO4` columns; a `dfCovs` with a raw `GENO` column instead
can be encoded on the fly with `oneHot = list(GENO = list(ref = 2))` and
`oneHotSep = ""`.

```{r}
dfresBS <- getForestDFSCM(
  dfCovs           = dfCovs,
  cdfCovsNames     = covnames,
  functionList     = list(paramFunction),
  functionListName = functionListName,
  noBaseThetas     = gen$noBaseThetas,
  dfParameters     = dfSamplesBS,
  dfRefRow         = dfRefRow
)
head(dfresBS[, c("COVNUM", "COVNAME", "GROUPNAME", "PARAMETER", "POINT", "Q1", "Q2")])
```

The model-derived reference from Section 4 can now be checked rather than taken
on trust. `POINT` is the parameter value for a row; `REFFUNC` is the reference
the row is divided by. For any covariate the model gives **no** effect on `CL` —
`AGE`, `CRCL`, `SEX`, `NCIL`, `FORM` are all absent from `run7`'s clearance —
the two should be identical, and the row should sit exactly at 1:

```{r}
subset(dfresBS, PARAMETER == "CL", c(COVNAME, POINT, REFFUNC))
```

They match to the last digit. Only `WT`, `FOOD = 0` and the genotypes move, which
are exactly the covariates `$PK` gives a clearance effect. Had the reference been
left at its data-derived default, every one of those rows would sit slightly off
1 — a displacement easy to read as a small covariate effect when it is only a
mismatched denominator.

Switching the uncertainty source is a one-argument change:

```{r}
dfresCOV <- getForestDFSCM(
  dfCovs           = dfCovs,
  cdfCovsNames     = covnames,
  functionList     = list(paramFunction),
  functionListName = functionListName,
  noBaseThetas     = gen$noBaseThetas,
  dfParameters     = dfSamplesCOV,
  dfRefRow         = dfRefRow
)
```

```{r}
forestPlot(dfresBS, parameters = "CL", groupNameLabels = covariateGroupNames)
```

# 6. Empirical Forest plots

An empirical Forest plot reads each covariate effect from the analysis data. The
covariate structure is a list of expressions that subset the data, and the
parameter function is the same one used above.

The parameter function needs the `GENO1`/`GENO3`/`GENO4` columns. This analysis
data set already contains them, but passing `oneHot = list(GENO = list(ref = 2))`
with `oneHotSep = ""` builds them from the raw `GENO` column when they are not
present, and is a harmless no-op when they are.

```{r}
lsExpr <- list(
  "NCI"  = expression(NCIL == 0),
  "NCI"  = expression(NCIL == 1),
  "FORM" = expression(FORM == 0),
  "FORM" = expression(FORM == 1),
  "FOOD" = expression(FOOD == 0),
  "FOOD" = expression(FOOD == 1),
  "GENO" = expression(GENO == 1),
  "GENO" = expression(GENO == 2),
  "GENO" = expression(GENO == 3),
  "GENO" = expression(GENO == 4),
  "WT"   = expression(WT < 70),
  "WT"   = expression(WT > 104),
  "SEX"  = expression(SEX == 1),
  "SEX"  = expression(SEX == 2)
)

covnamesEmp <- c(
  "NCI = 0", "NCI > 0", "Oral tablet", "FDC", "Fasted", "Fed",
  "2D6 UM", "2D6 EM", "2D6 IM", "2D6 PM",
  "WT < 70 kg", "WT > 104 kg", "Male", "Female"
)
stopifnot(length(covnamesEmp) == length(lsExpr))
```

`getForestDFemp()` needs the analysis data. Empirical plots are more expensive to
compute than parametric ones, so a subset keeps the example quick; `ncores > 1`
parallelises the calculation and `cstrPackages` names the packages the workers
need.

```{r}
dfDataEmp <- read.csv(dataFile) %>%
  distinct(ID, .keep_all = TRUE) %>%
  slice(1:300)

dfresEmp <- getForestDFemp(
  dfData             = dfDataEmp,
  covExpressionsList = lsExpr,
  cdfCovsNames       = covnamesEmp,
  functionList       = list(paramFunction),
  functionListName   = functionListName,
  noBaseThetas       = gen$noBaseThetas,
  dfParameters       = dfSamplesCOV,
  oneHot             = list(GENO = list(ref = 2)),
  oneHotSep          = "",
  ncores             = 1,
  cstrPackages       = "dplyr"
)
```

```{r}
forestPlot(dfresEmp, parameters = "CL", noVar = FALSE)
```

`noVar = FALSE` keeps the parameter-uncertainty interval. `noVar = TRUE` (the
default for empirical data) drops it and shows only the covariate effect.

# 7. Customise the plot

`forestPlot()` returns a `ggplot` object (or a list of them). Common arguments:

| Argument | Effect |
| --- | --- |
| `parameters` | which of `functionListName` to draw, and in what order |
| `groupNameLabels` | replace the covariate-group strip labels |
| `plotRelative` | `TRUE` (default) plots relative to the reference; `FALSE` plots the absolute value |
| `noVar` | drop the parameter-uncertainty interval |
| `xlim` | x-axis limits |
| `statisticsLabel` | prefix for the statistics-panel strip labels |
| `return` | `"plot"` (default) or `"plotList"` for the panel list |

```{r}
forestPlot(
  dfresBS,
  parameters      = c("CL", "FREL"),
  groupNameLabels = covariateGroupNames,
  statisticsLabel = "Estimate:",
  xlim            = c(0.5, 1.6)
)
```

# Where to next

* [Preparing Forest plot inputs](Part3-deep-dive-forest-plot-inputs.html) —
  all four inputs in full: the covariate structure, generating the parameter
  function from the control stream, the reference row, and keeping the last two
  consistent with each other.
* [R-coded models](Part3-deep-dive-r-coded-models.html) — Forest plots for models
  fitted outside NONMEM, with the plot data assembled by hand.
* [Time-to-event models](Part3-deep-dive-tte-models.html) — hazard ratios, event
  probabilities, and distribution reparameterisation.
