---
title: "1. Quick-Start: Forest Plots for NONMEM Models"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{1. Quick-Start: Forest Plots for NONMEM Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(dplyr)
  library(ggplot2)
})

theme_set(theme_bw(base_size = 14))
theme_update(plot.title = element_text(hjust = 0.5))

set.seed(865765)
```

# Introduction

This vignette is the shortest path from a fitted NONMEM model to a Forest plot. It
uses a PK model estimated with PsN and produces a parametric Forest plot of the
covariate effects on clearance, with the uncertainty taken from the NONMEM
covariance step.

A parametric Forest plot is built in five steps:

1. Describe the covariate values to display (`setupDfCovs()`).
2. Write the function that turns parameter estimates and covariate values into the
   quantities to plot.
3. Draw parameter vectors that carry the estimation uncertainty (`getSamples()`).
4. Compute the plot data (`getForestDFSCM()`).
5. Draw the plot (`forestPlot()`).

The [Walkthrough](Part2-walkthrough.html) covers the same steps in full, including
empirical Forest plots, alternative uncertainty sources, an explicit reference
subject, and the formatting arguments of `forestPlot()`.

## Which vignette do I want?

| If you want to… | Read |
| --- | --- |
| get a plot out of a NONMEM model quickly | this vignette |
| build a publication Forest plot end to end | [2. Walkthrough](Part2-walkthrough.html) |
| control how covariate values, references or rows are chosen | [3. Preparing Forest plot inputs](Part3-deep-dive-forest-plot-inputs.html) |
| plot AUC, half-life or another derived quantity | [3. Secondary parameters](Part3-deep-dive-secondary-parameters.html) |
| plot a model that is not a NONMEM `$PK` block | [3. R-coded models](Part3-deep-dive-r-coded-models.html) |
| plot hazard ratios or event probabilities | [3. Time-to-event models](Part3-deep-dive-tte-models.html) |

# 1. Describe the covariates

`setupDfCovs()` reads a data set, summarises each covariate, and returns
a data frame in which every row is one row of the Forest plot. Continuous
covariates are represented by their 5th and 95th percentiles; binary covariates
are expanded to one row per level.

```{r}
dataFile <- system.file("extdata", "SimVal/DAT-1-MI-PMX-2.csv", package = "PMXForest")
modFile <- system.file("extdata", "SimVal/run7.mod", package = "PMXForest")
dfData <- read.csv(dataFile)

covariates <- c("WT", "AGE", "CRCL", "SEX", "NCIL", "FORM", "FOOD", "RACEL2")

dfCovs <- setupDfCovs(data = dfData, covariates = covariates, idVar = "ID")
dfCovs
```

Cells that are inactive on a row hold the missing-value indicator `-99`; the
parameter function in the next step treats that as "no effect". Multi-level
categorical covariates (one-hot encoding, reference level) are covered in the
[Walkthrough](Part2-walkthrough.html).

# 2. Generate the parameter function

The parameter function turns the estimates and one row of covariate values into
the quantities to plot. Rather than retyping the model's algebra in R,
`createParamFunction()` translates the `$PK` block of the control stream.

```{r}
gen <- createParamFunction(
  modFile,
  parameters = c("CL", "FREL", "V"),
  covRef     = list(GENO1 = 0, GENO3 = 0),
  secondary  = list(AUC = "80 / (CL / FREL)"),
  quiet      = TRUE
)
```

`covRef` pins down two genotype dummies the control stream does not settle on its
own -- without it the generator proposes a value and warns. `secondary` adds
`AUC` after an 80 mg dose, which is not in `$PK` because the dose is not.

The result is **source text, not a function**. Read it against your control
stream before you run it: the missing-value handling is hoisted into an
annotated preamble, and everything below is a plain transliteration of `$PK`.

```{r}
cat(gen$code, sep = "\n")
```

Once it matches the model, evaluate it:

```{r}
paramFunction <- eval(parse(text = paste(gen$code, collapse = "\n")))
```

`gen` also carries the parameter names and the THETA count, so neither has to be
typed in and kept up to date by hand.

If the translation is not right for what you need, the source text is the
starting point rather than a dead end —
[Preparing Forest plot inputs, Section 2.4](Part3-deep-dive-forest-plot-inputs.html)
shows how to adjust it without losing the parts that were right.

::: {.callout-note}
If the model was run with a `$TABLE` holding the parameters and the covariates,
`verifyParamFunction()` checks the translation against the values NONMEM itself
wrote. No `$TABLE` is bundled here, so it is not shown; see the
[Walkthrough](Part2-walkthrough.html) and
[Preparing Forest plot inputs](Part3-deep-dive-forest-plot-inputs.html).
:::

# 3. Sample the parameter uncertainty

`getSamples()` reads the `.ext` file for the final estimates and the `.cov` file
for the covariance matrix, then draws `n` parameter vectors from the implied
multivariate normal distribution.

```{r}
extFile <- system.file("extdata", "SimVal/run7.ext", package = "PMXForest")
covFile <- system.file("extdata", "SimVal/run7.cov", package = "PMXForest")

dfSamples <- getSamples(covFile, extFile = extFile, n = 200)
```

# 4. Compute the Forest plot data

`getForestDFSCM()` evaluates `paramFunction` for every row of `dfCovs` and every
parameter vector in `dfSamples`, then summarises the results into the columns
`forestPlot()` needs. `noBaseThetas` is the number of THETAs in the model.

With `cdfCovsNames` left at its default, each row is labelled from its own
covariate values (`WT=58.5`, `SEX=2`, and so on). The
[Walkthrough](Part2-walkthrough.html) shows how to supply publication-quality row
labels and covariate group names.

```{r}
dfres <- getForestDFSCM(
  dfCovs           = dfCovs,
  functionList     = list(paramFunction),
  functionListName = gen$functionListName,
  noBaseThetas     = gen$noBaseThetas,
  dfParameters     = dfSamples
)
```

# 5. Draw the plot

```{r}
forestPlot(dfres, parameters = "CL")
```

The dotted line marks the reference. No `dfRefRow` was supplied, so the reference
is the row with every covariate set to `-99`. With a generated parameter
function that is the right reference by construction: `-99` sends each covariate
to the value the preamble read out of `$PK`, so the reference *is* the model's
own reference subject rather than something computed separately that might
disagree with it. Supplying a reference taken from the data instead — a median
weight, the most common formulation — is a deliberate choice, and one the
[Walkthrough](Part2-walkthrough.html) makes explicitly.

The points show the covariate effect on `CL` relative to that reference, and the
bars its 90% confidence interval. `WT` is fixed in this model, so its interval
has zero width.

::: {.callout-note}
**Next steps.** The [Walkthrough](Part2-walkthrough.html) adds an explicit
reference subject, empirical Forest plots, bootstrap and SIR uncertainty, and the
formatting arguments of `forestPlot()`. The deep-dive vignettes cover
[preparing Forest plot inputs](Part3-deep-dive-forest-plot-inputs.html),
[R-coded models](Part3-deep-dive-r-coded-models.html), and
[time-to-event models](Part3-deep-dive-tte-models.html).
:::
