---
title: "3. Deep Dive: How the NONMEM Model is Parsed"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: How the NONMEM Model is Parsed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
.callout-warning {
  background-color: #fff3cd;
  border-left: 4px solid #ffc107;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

```{r setup}
library(PMXForest)
```

# Introduction

[createParamFunction()](../reference/createParamFunction.html) reads a `$PK`
block and writes the R that a Forest plot is computed from. That makes it the
most consequential function in the package: a mistake in it does not produce an
error, it produces a plot.

This vignette says what the translation promises, so you can judge whether it
holds. It is not a tour of the source. It is the set of claims the generator
makes, each one demonstrated on a model small enough to check by eye.

Every model here is written in the vignette itself, one per idiom, because a
model that already works confirms the parser rather than tests the explanation.

```{r}
# Write a throwaway control stream and return its path.
mod <- function(...) {
  f <- tempfile(fileext = ".mod")
  writeLines(c(...), f)
  f
}

# Show what a call refuses, without stopping the vignette.
refusal <- function(expr) {
  msg <- tryCatch(
    {
      expr
      "no refusal"
    },
    error = function(e) conditionMessage(e)
  )
  cat(strwrap(msg, 76), sep = "\n")
}
```

# 1. What the grammar accepts

Assignments, `IF` constructs, and closed-form arithmetic. Every `THETA(n)`
becomes `thetas[n]`, every `ETA(n)` becomes 0, and the covariates are lifted
into an annotated preamble.

```{r}
m <- mod(
  "$PROBLEM grammar", "$INPUT ID TIME DV AMT WT",
  "$DATA d.csv IGNORE=@", "$PK",
  "TVCL = THETA(1)",
  "CL   = TVCL * (WT/75)**THETA(2) * EXP(ETA(1))",
  "$THETA (0,7) (0,0.75)"
)
gen <- createParamFunction(m, parameters = "CL", quiet = TRUE)
cat(gen$code, sep = "\n")
```

Two things to notice. `**` became `^`, and `ETA(1)` became `exp(0)` rather than
being dropped — the generated function returns *typical* values, and it says so
in a comment on the line where it happened.

## 1.1 Why a faithful translation is not the obvious one

Fortran and R disagree about `MOD()`, in two independent ways. `%%` binds
tighter than `*` and `/` in R while `MOD()` is a call, so the obvious
transliteration of `MOD(A*B, C)` gives `A * (B %% C)`. And Fortran's `MOD()`
truncates towards zero where R's `%%` floors, so the two disagree in sign for a
negative first argument.

```{r}
m <- mod(
  "$PROBLEM mod", "$INPUT ID TIME DV AMT OCC",
  "$DATA d.csv IGNORE=@", "$PK",
  "CYCLE = MOD(OCC*2, 3)",
  "CL    = THETA(1) * (1 + CYCLE)",
  "$THETA (0,7)"
)
gen <- createParamFunction(m, parameters = "CL", covRef = list(OCC = 0), quiet = TRUE)
grep("CYCLE", gen$code, value = TRUE)
```

`a - b * trunc(a / b)` is correct for either sign and gets its parentheses from
ordinary precedence.

```{r}
f <- eval(parse(text = paste(gen$code, collapse = "\n")))
# Fortran MOD(-7, 3) is -1. R's -7 %% 3 is 2.
c(fortran = -7 - 3 * trunc(-7 / 3), r_modulo = -7 %% 3)
```

# 2. Folding the ETAs

The generated function returns typical values, so every `ETA(n)` goes to 0. The
generator also records **which** eta belongs to which parameter, in `etaMap`,
and [verifyParamFunction()](../reference/verifyParamFunction.html) uses that to
recover a typical value from a table of individual ones by dividing out
`exp(eta)`.

That division is only valid when the eta separates. Two idioms separate, and
they are the two in common use:

```{r}
m <- mod(
  "$PROBLEM etas", "$INPUT ID TIME DV AMT WT",
  "$DATA d.csv IGNORE=@", "$PK",
  "TVCL = THETA(1)", "TVV = THETA(2)",
  "MU_2 = LOG(TVV)",
  "CL   = TVCL * EXP(ETA(1))", # classic
  "V    = EXP(MU_2 + ETA(2))", # MU-referenced, as IMP and SAEM write it
  "$THETA (0,7) (0,3)"
)
createParamFunction(m, parameters = c("CL", "V"), quiet = TRUE)$etaMap
```

They separate for the same reason: `EXP(a + eta)` is `EXP(a) * EXP(eta)`.

## 2.1 When the eta does not separate

An eta that is scaled, transformed, or shared does not come out of the
expression by division, and no entry is claimed for it.

```{r}
m <- mod(
  "$PROBLEM inseparable", "$INPUT ID TIME DV AMT WT",
  "$DATA d.csv IGNORE=@", "$PK",
  "TVCL = THETA(1)", "TVKA = THETA(2)", "TVF = THETA(3)",
  "BXPAR = THETA(4)",
  "CL   = TVCL * EXP(ETA(1))",
  # a shared eta, scaled by a theta
  "KA   = TVKA * EXP(ETA(1)*(1 + THETA(5)))",
  # Box-Cox transformed
  "BSV  = (EXP(ETA(2))**BXPAR - 1)/BXPAR",
  "F1   = TVF * EXP(BSV)",
  "$THETA (0,7) (0,1) (0,1) (-2,0.5) (-1,0.9)"
)
createParamFunction(m, parameters = c("CL", "KA", "F1"), quiet = TRUE)$etaMap
```

Only `CL` is there. `KA` and `F1` are translated perfectly well — they are
simply not parameters whose typical value can be recovered from a tabled
individual value by division, so `verifyParamFunction()` will say it cannot
check them rather than check them wrongly.

The third way an eta fails to separate is the one that is easiest to miss,
because nothing in the expression looks wrong. Inter-occasion variability is
normally assigned to a name in an occasion block and then added inside the
exponent, so the expression contains exactly one `ETA()` and still carries two
sources of randomness:

```{r}
m <- mod(
  "$PROBLEM iov", "$INPUT ID TIME DV AMT OCC",
  "$DATA d.csv IGNORE=@", "$PK",
  "TVCL = THETA(1)", "TVV = THETA(2)",
  "IOV  = 0",
  "IF(OCC.EQ.1) IOV = ETA(3)",
  "IF(OCC.EQ.2) IOV = ETA(4)",
  "CL   = TVCL * EXP(ETA(1) + IOV)", # two sources, one ETA() node
  "V    = TVV  * EXP(ETA(2))",
  "$THETA (0,7) (0,3)"
)
createParamFunction(m,
  parameters = c("CL", "V"),
  covRef = list(OCC = 1), quiet = TRUE
)$etaMap
```

`V` is claimed and `CL` is not, which is the answer we want: dividing `CL`'s
tabled individual value by `exp(eta1)` would leave `exp(IOV)` behind, and
`verifyParamFunction()` would report a mismatch on arithmetic that was never
comparable in the first place. The generator resolves each name in the exponent
back to what assigned it, so a symbol that carries randomness disqualifies the
term just as a visible `ETA()` would.

The older way of coding the same thing puts the occasion etas in the exponent
directly, switched by an indicator. Three `ETA()`s, so the one-eta rule declines
it for the same reason:

```{r}
m <- mod(
  "$PROBLEM iov classic", "$INPUT ID TIME DV AMT WT OCCOBS",
  "$DATA d.csv IGNORE=@", "$PK",
  "TVCL = THETA(1)", "TVV = THETA(2)",
  "OCC = 0",
  "IF(OCCOBS.EQ.2) OCC = 1",
  "CL = TVCL*EXP(ETA(1) + ETA(2)*(1-OCC) + ETA(3)*OCC)",
  "V  = TVV*EXP(ETA(4))",
  "$THETA (0,7) (0,3)"
)
out <- createParamFunction(m, parameters = c("CL", "V"), quiet = TRUE)
out$etaMap
```

`CL` has no typical value recoverable *by division*, but it does still have a
typical value: with every eta at 0 the whole IOV term vanishes and `CL` is
`TVCL`. That is what `tvMap` records, so `verifyParamFunction()` can compare
`CL` against a `TVCL` column in a `$TABLE` even though `etaMap` is empty for it:

```{r}
out$tvMap
grep("^  CL <-", out$code, value = TRUE)
```

Because the term folds away entirely, the occasion variable is not a covariate
of the typical value and needs no reference:

```{r}
"OCC" %in% names(out$covRef)
```

<div class="callout-note">
`TV` in front of a parameter name is a convention, not a rule. `tvMap` records
the symbol the control stream itself treats as each parameter's typical value,
and `verifyParamFunction()` uses that rather than guessing from the name — a
model can perfectly well assign `TVD1` as a fraction that `D1` is computed
from.
</div>

```{r}
m <- mod(
  "$PROBLEM tvmap", "$INPUT ID TIME DV AMT",
  "$DATA d.csv IGNORE=@", "$PK",
  "TVMAT = THETA(1)", "TVD1 = THETA(2)",
  "MAT  = TVMAT * EXP(ETA(1))", # TVMAT really is MAT's typical value
  "D1   = MAT * (1 - TVD1)", # TVD1 is a fraction, not D1's typical value
  "$THETA (0,2) (0,0.8)"
)
out <- createParamFunction(m, parameters = c("MAT", "D1"), quiet = TRUE)
out$tvMap
"D1" %in% names(out$tvMap)
```

# 3. Where a reference value comes from

A parameter function has to return something when a covariate is inactive, and
what it returns is the model's own reference. Four rules are tried in order,
and each records where it got its answer.

```{r}
m <- mod(
  "$PROBLEM refs", "$INPUT ID TIME DV AMT WT AGE FOOD SEX",
  "$DATA d.csv IGNORE=@", "$PK",
  # rule 1: explicit missing-value handling
  "IF(AGE.EQ.-99) AGE = 50",
  # rule 2: the branch PsN's scm marks "; Most common"
  "IF(FOOD.EQ.1) CLFOOD = 1  ; Most common",
  "IF(FOOD.EQ.0) CLFOOD = (1 + THETA(3))",
  # rule 2b: the branch assigning the identity value
  "IF(SEX.EQ.1) CLSEX = 1",
  "IF(SEX.EQ.2) CLSEX = (1 + THETA(4))",
  # rule 3: the normalisation constant
  "CLWT = (WT/75)**THETA(2)",
  "CL   = THETA(1)*CLWT*CLFOOD*CLSEX*(1 + THETA(5)*(AGE - 50))",
  "$THETA (0,7) (0,0.75) (-1,0.2) (-1,0.1) (-1,0.01)"
)
out <- createParamFunction(m, parameters = "CL", quiet = TRUE)
for (cv in names(out$covRef)) {
  cat(sprintf(
    "%-5s %-5s confident=%-5s %s\n", cv, format(out$covRef[[cv]]$value),
    out$covRef[[cv]]$confident, out$covRef[[cv]]$source
  ))
}
```

## 3.1 The fifth rule proposes, and says so

Where no rule fits, a level that no `IF()` tests is proposed and the call warns.
`confident` is `FALSE`, and that flag is the only signal you get.

```{r}
m <- mod(
  "$PROBLEM proposed", "$INPUT ID TIME DV AMT GENO1",
  "$DATA d.csv IGNORE=@", "$PK",
  "CLG = 1",
  "IF(GENO1.EQ.1) CLG = (1 + THETA(2))",
  "CL  = THETA(1)*CLG",
  "$THETA (0,7) (-1,0.3)"
)
out <- suppressWarnings(createParamFunction(m, parameters = "CL", quiet = TRUE))
out$covRef$GENO1[c("value", "confident", "source")]
```

## 3.2 A rule can read a structure backwards

<div class="callout-warning">
This is the failure mode worth understanding, because nothing downstream
catches it. A wrong reference displaces every row of a Forest plot, and
[verifyParamFunction()](../reference/verifyParamFunction.html) takes covariate
values **from the `$TABLE`**, so the fallback path a reference lives on is never
exercised by a verification run.
</div>

Rule 2b reads "the branch assigning a bare `0` or `1`" as the reference
category. That is right for the SCM shape it was written for — and backwards
when the variable also has an unconditional default:

```{r}
m <- mod(
  "$PROBLEM indicator", "$INPUT ID TIME DV AMT COV",
  "$DATA d.csv IGNORE=@", "$PK",
  "IND  = 0", # the reference state
  "IF(COV.EQ.2) IND = 1", # the departure from it
  "TVCL = THETA(1) + IND*THETA(2)",
  "CL   = TVCL",
  "$THETA (0,7) (-1,0.5)"
)
out <- suppressWarnings(createParamFunction(m, parameters = "CL", quiet = TRUE))
out$covRef$COV[c("value", "confident", "source")]
```

`IND` is an indicator, so its identity is `0`, and the unconditional assignment
is the reference state. Reading the `IF` as the reference names the *treated*
level. The rule is skipped for a variable that is also assigned unconditionally,
so this falls through to the proposing rule — which is the honest answer, since
the right value is not recoverable from `$PK` at all.

Read the preamble against the control stream before you trust it. That is the
check the generator cannot do for you.

# 4. What it refuses, and why

Refusing beats guessing when the guess would look plausible.

```{r}
base <- c("$PROBLEM refuse", "$INPUT ID TIME DV AMT WT", "$DATA d.csv IGNORE=@")

# a compartment amount is not available to a parameter function
refusal(createParamFunction(mod(
  base, "$PK", "CONC = A(2)/THETA(1)", "CL = THETA(1)*CONC", "$THETA (0,7)"
), quiet = TRUE))
```

```{r}
# a symbol read before anything assigns it
refusal(createParamFunction(mod(
  base, "$PK", "CL = TVCL*(WT/75)", "TVCL = THETA(1)", "$THETA (0,7)"
), parameters = "CL", quiet = TRUE))
```

NONMEM neither initialises `$PK` variables nor clears them between data
records, so such a model reads whatever the previous subject left behind. A
parameter function is evaluated one row at a time and cannot reproduce that.

```{r}
# a FREM model
refusal(createParamFunction(mod(
  "$PROBLEM frem", "$INPUT ID TIME DV AMT WT FREMTYPE",
  "$DATA d.csv IGNORE=@", "$PK", "CL = THETA(1)", "$THETA (0,7)"
), quiet = TRUE))
```

A FREM model's `$PK` translates faithfully, and that is the trap — its
covariate effects are in `$OMEGA`, so the function produced describes none of
the covariates the model was built for.

## 4.1 Verbatim FORTRAN, which you can override

Verbatim code can define variables the rest of `$PK` reads, and nothing here
can tell that apart from a solver directive that touches no parameter. So it is
refused by default and the decision is yours:

```{r}
verb <- mod(
  base, "$PK",
  '  "FIRST',
  '  " USE PRDATA, ONLY: MXSTP01',
  '  " MXSTP01=2147483647',
  "CL = THETA(1)*(WT/75)**0.75", "$THETA (0,7)"
)
refusal(createParamFunction(verb, parameters = "CL", quiet = TRUE))
```

```{r}
out <- createParamFunction(verb, parameters = "CL", ignoreVerbatim = TRUE, quiet = TRUE)
eval(parse(text = paste(out$code, collapse = "\n")))(thetas = 7, df = data.frame(WT = 75))
```

# 5. What you have to supply

## 5.1 A symbol NONMEM supplies rather than `$INPUT`

A `$MIX` model reads `MIXNUM`, the subpopulation being computed for, and
usually `MIXEST`, the one the individual was classified into. `$PK` never
assigns them and `$INPUT` never declares them, so only you can say what value
to hold them at.

```{r}
mix <- mod(
  "$PROBLEM mixture", "$INPUT ID TIME DV AMT WT",
  "$DATA d.csv IGNORE=@", "$PK",
  "TVCL = THETA(1)",
  "IF(MIXNUM.EQ.2) TVCL = THETA(2)",
  "IF(MIXEST.EQ.2) TVCL = TVCL*1.1",
  "CL = TVCL*(WT/75)**0.75",
  "$THETA (0,7) (0,3)", "$MIX"
)
refusal(createParamFunction(mix, parameters = "CL", quiet = TRUE))
```

Everything it cannot account for is named at once, with a `covRef` call you can
paste in. Pinned, the symbol becomes an ordinary covariate — so giving `MIXNUM`
its own column in `dfCovs` plots each subpopulation in turn.

```{r}
out <- createParamFunction(mix,
  parameters = "CL",
  covRef = list(MIXNUM = 1, MIXEST = 1), quiet = TRUE
)
f <- eval(parse(text = paste(out$code, collapse = "\n")))
c(
  subpop1 = f(thetas = c(7, 3), df = data.frame(WT = 75, MIXNUM = 1, MIXEST = 1))$CL,
  subpop2 = f(thetas = c(7, 3), df = data.frame(WT = 75, MIXNUM = 2, MIXEST = 1))$CL
)
```

`NEWIND` is the case that cannot be pinned meaningfully: it changes within an
individual, so a single value describes one kind of record rather than a
subject.

## 5.2 The model's own variance matrix

A model that builds a correlation by hand reads `OMEGA(i,j)`. Those are
constants once the model is estimated, so they fold to literals — but they live
in the `.ext` file, not the control stream.

```{r}
om <- mod(
  "$PROBLEM omega", "$INPUT ID TIME DV AMT WT",
  "$DATA d.csv IGNORE=@", "$PK",
  "SD = SQRT(OMEGA(2,2))",
  "CL = THETA(1)*(1 + SD)",
  "$THETA (0,7)"
)
refusal(createParamFunction(om, parameters = "CL", quiet = TRUE))
```

```{r}
extFile <- system.file("extdata", "SimVal/run7.ext", package = "PMXForest")
out <- createParamFunction(om, parameters = "CL", extFile = extFile, quiet = TRUE)
grep("SD <-", out$code, value = TRUE)
```

The `.ext` holds the lower triangle and the matrix is symmetric, so either
index order resolves.

# 6. Asking for less

`parameters` prunes the source as well as the return list. Only the statements
the requested parameters can reach are emitted, and the covariate set narrows
with them — so a reference never has to be justified for a covariate that
cannot affect the answer.

```{r}
modFile <- system.file("extdata", "SimVal/run7.mod", package = "PMXForest")
shape <- function(p) {
  o <- suppressWarnings(createParamFunction(modFile, parameters = p, quiet = TRUE))
  c(
    returns = length(o$functionListName),
    statements = sum(grepl("^  [A-Z]", o$code)),
    covariates = length(o$covRef)
  )
}
rbind(all = shape(NULL), CL = shape("CL"), V = shape("V"))
```

A `secondary` block is spliced in afterwards and sees every structural
parameter by name, so anything it reads is kept even when you did not ask for
it.

```{r}
out <- suppressWarnings(createParamFunction(modFile,
  parameters = c("CL", "V"),
  secondary = list(AUC = "80 / (CL / FREL)"), quiet = TRUE
))
c(
  returned = paste(out$functionListName, collapse = ", "),
  FREL_kept = any(grepl("FREL", out$code))
)
```

# 7. `$PRED` models

Everything above is written against `$PK`, but `$PRED` is the same abbreviated
code and is read the same way. What differs is what else is in the block.

A `$PK` block assigns parameters and stops. A `$PRED` block computes the
prediction *and* the residual error in the same lines, so it also contains `Y`
and `EPS()`. There is no structural signal separating "these are parameters"
from "this is the prediction machinery" — which is why `parameters` is required
rather than optional here:

```{r}
pred <- mod(
  "$PROBLEM hand-coded one-compartment",
  "$INPUT ID TIME DV AMT WT",
  "$DATA d.csv IGNORE=@", "$PRED",
  "TVCL = THETA(1)*(WT/70)**THETA(3)",
  "TVV  = THETA(2)",
  "CL   = TVCL*EXP(ETA(1))",
  "V    = TVV*EXP(ETA(2))",
  "F    = AMT/V*EXP(-CL/V*TIME)",
  "Y    = F + F*EPS(1)",
  "$THETA (0,7) (0,30) (0,0.75)"
)
refusal(createParamFunction(pred, quiet = TRUE))
```

Name them and it behaves exactly as the `$PK` equivalent would:

```{r}
out <- createParamFunction(pred, parameters = c("CL", "V"), quiet = TRUE)
out$etaMap
out$tvMap
eval(parse(text = paste(out$code, collapse = "\n")))(
  thetas = c(7, 30, 0.75), df = data.frame(WT = 70)
)
```

`EPS(n)` folds to 0 exactly as `ETA(n)` does, so `Y` is an assignment like any
other — ask for it and you get the typical prediction, with the residual error
gone:

```{r}
y <- createParamFunction(pred,
  parameters = "Y", covRef = list(AMT = 100, TIME = 0), quiet = TRUE
)
grep("^  Y <-", y$code, value = TRUE)
```

<div class="callout-warning">
`ERR(n)` is ambiguous in NM-TRAN's own definition: it means `EPS(n)` for
population data and `ETA(n)` for single-subject data. Both fold to 0, so the
typical value is right either way — but the generator cannot tell which it is,
so a parameter whose variability is written `ERR(n)` never gets an `etaMap`
entry, and `verifyParamFunction()` will decline to check it. Write `ETA(n)` if
you want it checked.
</div>

# 8. The gaps

Three, stated rather than hidden.

**Definite assignment is weaker than it looks.** A symbol assigned in *some*
branch and read afterwards is accepted, even though no branch may have run. The
strict rule — assigned on every path — rejected the package's own reference
model, which is ordinary PsN `scm` output:

```
IF(FORM.EQ.1.0000E+00) FRELFORM = 1  ; Most common
IF(FORM.EQ.0.0000E+00) FRELFORM = ( 1 + THETA(12))
```

Two one-line `IF`s with no `ELSE`, exhaustive only because `FORM` is 0 or 1 — a
fact no static analysis can see, and structurally identical to the genuinely
non-exhaustive case the check existed to catch. The rule was weakened to accept
both rather than special-case the example away.

**A single-subject `$PRED` model loses its `etaMap`.** As above: `ERR(n)` could
be either an eta or an eps, and guessing would make
[verifyParamFunction()](../reference/verifyParamFunction.html) divide by
something that is not there. The cost is that such a model is translated but not
checkable.

**A reference value cannot be verified by anything here.**
[verifyParamFunction()](../reference/verifyParamFunction.html) checks the
algebra against a `$TABLE`, using covariate values from that table — so the
reference, which only applies when a covariate is *absent*, is never on the
path it tests. The `confident` flag and the warning are the whole of the
defence.

<div class="callout-note">
Which is why the generated source is returned as **text**. Read the preamble
against the control stream, confirm each reference, and evaluate it when it
matches. See
[Preparing Forest plot inputs](Part3-deep-dive-forest-plot-inputs.html) for
adjusting the result when it is not quite what a particular plot needs.
</div>
