---
title: "3. Deep Dive: Preparing Forest Plot Inputs"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Preparing Forest Plot Inputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
.callout-warning {
  background-color: #fff3cd;
  border-left: 4px solid #ffc107;
  color: #664d03;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(dplyr)
})

set.seed(865765)
```

# Introduction

`getForestDFSCM()` takes four inputs. The [Quick start](Part1-quick-start.html)
and the [Walkthrough](Part2-walkthrough.html) build all four with the
convenience functions, which is the right way to work. This vignette is for the
times that is not enough: when you need to know what a convenience function
decided on your behalf, override one step of it, or build an input the wrappers
do not cover.

So the table below reads in both directions. Left to right, it is what each
input is and which one call produces it. Right to left, it is what that call
does underneath — the function to reach for when you want to take over.

| Input | What it is | One call | What that call does underneath |
| --- | --- | --- | --- |
| `dfCovs` | one row per Forest plot row, inactive cells at `-99` | `setupDfCovs()` | `getCovStats()` picks the values to plot, then `createInputForestData()` lays them out as rows |
| `functionList` | the parameter function: typical values from thetas and covariates | `createParamFunction()` | `nmParsePK()` parses `$PK`, `nmCovRef()` finds a reference for each covariate, then the tree is emitted as R source |
| `dfRefRow` | the reference subject the other rows are expressed relative to | `setupDfRefRow()` | `refResolve()` resolves one reference per covariate (`refValues()` from the data, the `$PK` for `"model"`), shaped into a single row matching `dfCovs` |
| `dfParameters` | parameter uncertainty samples | `getSamples()` | this is the primitive — it reads the `.cov`, bootstrap or SIR file directly |
| `covExpressionsList` | which subjects each empirical row summarises | `setupCovExpressionsList()` | applies the same level and quantile rules as `getCovStats()`, but **implements them separately** rather than calling it, because the two return different shapes |

The empirical entry point `getForestDFemp()` takes a `covExpressionsList` in place
of `dfCovs`; that is covered in Section 7.

There is a third way to read the table, and it matters more than either. A
convenience function encodes one reasonable answer to "what should this input
look like?" — the observed percentiles, one row per level, the reference the
control stream implies. When that answer does not fit the question you are
asking, the fix is usually **not** to abandon the function and build the input
by hand. It is to take what the function produced and change the part that is
wrong, in script rather than by retyping.

That keeps the provenance: what came from the data or the model is still
derived, the deviation is a visible line of code that a reviewer can see and
question, and re-running after the model or the data changes reproduces both
the derivation and the deviation. Sections 1 and 2 each end with a worked
example.

Two of those rows are worth pausing on. `setupDfCovs()` is a genuine composition
— you can call `getCovStats()` and `createInputForestData()` yourself and get the
same object, which Section 1 does step by step. `setupCovExpressionsList()` is
not: it duplicates the logic. If you change how one of them chooses levels or
quantiles, the other does not follow.

The data throughout is the PK data file used in the
[Walkthrough](Part2-walkthrough.html), read as it comes rather than filtered
through the model's `IGNORE` statements, and the model is the `run7` control
stream bundled with the package.

```{r}
dataFile <- system.file("extdata", "SimVal/DAT-1-MI-PMX-2.csv", package = "PMXForest")
modFile <- system.file("extdata", "SimVal/run7.mod", package = "PMXForest")
dfData <- read.csv(dataFile)
```

# 1. Covariate values: `dfCovs`

## `getCovStats()`

`getCovStats()` returns a named list, one element per covariate. The element type
depends on the number of unique values:

* **continuous** (more than `minLevels` unique values) — a two-element numeric
  vector of quantiles, by default the 5th and 95th percentiles;
* **binary** — the two levels, sorted alphanumerically;
* **multi-level categorical** — a nested list of one-hot vectors. The reference
  level (the lowest level, or the one named in `catRef`) is the all-zero row.

```{r}
covStats <- getCovStats(
  data       = dfData,
  covariates = c("WT", "AGE", "SEX", "GENO"),
  idVar      = "ID"
)
covStats
```

Statistics are computed on the first record per `idVar`, so subjects with dense
sampling do not pull the quantiles.

```{r}
nrow(dfData) # all records
nrow(distinct(dfData, ID)) # subjects
```

`minLevels` (default 10) is the threshold between categorical and continuous. A
covariate coded with a small number of distinct numeric values — a dose level, an
ordinal score — is treated as categorical unless `minLevels` is lowered.

```{r}
## GENO has 4 levels: categorical by default
getCovStats(dfData, "GENO", idVar = "ID")

## Force GENO to be treated as continuous
getCovStats(dfData, "GENO", idVar = "ID", minLevels = 3)
```

`probs` sets the quantiles for continuous covariates; `nsig` sets the rounding.

```{r}
getCovStats(dfData, "CRCL", idVar = "ID", probs = c(0.1, 0.9), nsig = 4)
```

## `catRef` and `sep`

By default the lowest level of a multi-level categorical is the reference (the
all-zero row) and the one-hot columns are named `<covariate>_<level>`.
`catRef` overrides the reference level per covariate; `sep` changes the
separator. (`catRef` replaces `refLevels`, which still works but is deprecated.)

```{r}
## GENO reference is level 1 by default -> columns GENO_2, GENO_3, GENO_4
getCovStats(dfData, "GENO", idVar = "ID")

## Make level 2 the reference -> columns GENO_1, GENO_3, GENO_4
getCovStats(dfData, "GENO", idVar = "ID", catRef = list(GENO = 2))
```

Use `catRef` when the model's reference category is not the lowest level —
otherwise the one-hot columns will not line up with the model's coefficients. The
`"_"` separator matches NONMEM FREM data sets and `PMXFrem::createFREMModel()`;
change it only if you have a reason to.

::: {.callout-note}
Binary levels are returned in alphanumeric order (`c(0, 1)`), not order of
appearance in the data. Forest plot row order is therefore reproducible across
data sets.
:::

## `createInputForestData()`

`createInputForestData()` turns a covariate list into the row structure. It is
what you call directly when the covariate values are known literals rather than
derived from data.

```{r}
dfCovsManual <- createInputForestData(
  list(
    "WT"   = c(60, 100),
    "SEX"  = c(1, 2),
    "GENO" = list("GENO_2" = c(0, 1, 0), "GENO_3" = c(0, 0, 1), "GENO_4" = c(0, 0, 0))
  ),
  iMiss = -99
)
dfCovsManual
```

The `COVARIATEGROUPS` column records which covariate each row belongs to;
`getForestDFSCM()` uses it to group and space the rows on the y-axis, and drops it
before calling the parameter function.

A `getCovStats()` result can be passed straight in:

```{r}
createInputForestData(covStats, iMiss = -99)
```

## `setupDfCovs()`

`setupDfCovs()` runs `getCovStats()` and `createInputForestData()` together and
adds control over how inactive cells are filled.

```{r}
covariates <- c("WT", "AGE", "CRCL", "SEX", "NCIL", "GENO")

dfCovs <- setupDfCovs(data = dfData, covariates = covariates, idVar = "ID")
dfCovs
```

Covariates listed in `additionalCovs` get their own rows, but on every *other*
row they sit at their reference rather than at `-99`. This is for covariates that
must hold a plausible value whenever they are not the covariate under study — for
example one carried through from a full random-effects (FREM) model.

```{r}
setupDfCovs(
  data           = dfData,
  covariates     = c("WT", "AGE", "SEX"),
  additionalCovs = "CRCL",
  idVar          = "ID"
)
```

On the `WT`, `AGE`, and `SEX` rows, `CRCL` holds its median instead of `-99`.

With `useMissVal = FALSE`, inactive *primary* covariates are also set to their
reference, so the parameter function never sees `-99`.

```{r}
setupDfCovs(
  data       = dfData,
  covariates = c("WT", "AGE", "SEX"),
  useMissVal = FALSE,
  idVar      = "ID"
)
```

## Adjusting what `setupDfCovs()` produced

`setupDfCovs()` gives one row per level and the 5th/95th percentiles of the
observed data. Two things it will not do: use a value the study did not observe,
and show a *combination* of covariates, since it varies one at a time.

Both are ordinary requirements — a label may specify weight bands rather than
percentiles, and the clinical question is often "what about a light poor
metaboliser?", which no single-covariate row answers. Start from the generated
structure and edit it.

```{r}
dfCovsAdj <- setupDfCovs(dfData,
  covariates = c("WT", "GENO"),
  catRef = list(GENO = 2), sep = "", idVar = "ID"
)
dfCovsAdj
```

A combination row is built from any existing row, so it is guaranteed to have
the right columns in the right order. Blank it to `missVal`, then set only the
cells the row is about:

```{r}
combo <- dfCovsAdj[1, ]
combo[setdiff(names(combo), "COVARIATEGROUPS")] <- -99
combo$WT <- 60
combo$GENO1 <- 0
combo$GENO3 <- 0
combo$GENO4 <- 1
combo$COVARIATEGROUPS <- "Combination"

dfCovsAdj <- rbind(dfCovsAdj, combo)
tail(dfCovsAdj, 2)
```

Note that all three genotype columns are set, not just `GENO4`. A one-hot
covariate is only inactive when every one of its columns holds `missVal`;
leaving `GENO1` and `GENO3` at `-99` while `GENO4` is `1` is a contradiction the
parameter function would resolve by falling back to the reference for the two it
cannot read.

Row labels are supplied separately and are not regenerated, so extend them in
the same step and keep the `stopifnot()` guard:

```{r}
covnamesAdj <- c(
  paste0("WT ", dfCovsAdj$WT[1:2], " kg"),
  "2D6 UM", "2D6 EM", "2D6 IM", "2D6 PM",
  "60 kg, 2D6 PM"
)
stopifnot(length(covnamesAdj) == nrow(dfCovsAdj))
```

The same approach covers replacing a percentile with a protocol value
(`dfCovsAdj$WT[1:2] <- c(60, 100)`) or dropping a row you do not want to show
(`dfCovsAdj[-3, ]`). What it does not cover is a *different shape* of input —
for that, `createInputForestData()` builds one from literal values directly, as
above.

## `oneHotEncode()`

`getCovStats()` produces one-hot vectors from a *summary* of the data.
`oneHotEncode()` does the same job on a *data frame*: it adds
`<covariate><sep><level>` dummy columns to whatever data frame you give it. This
is what the `oneHot` argument of `getForestDFSCM()` and `getForestDFemp()` calls
internally, and it follows the same convention as NONMEM FREM data sets.

```{r}
raw <- data.frame(
  ID   = 1:6,
  GENO = c(1, 2, 3, 4, 2, -99),
  RACE = c(1, 1, 2, 3, 2, 1)
)

## Lowest level as reference; missing rows kept as missVal
oneHotEncode(raw, spec = c("GENO", "RACE"))
```

The reference level is set the same way as in `getCovStats()`:

```{r}
oneHotEncode(raw, spec = list(GENO = list(ref = 2)), dropOriginal = TRUE)
```

| Argument | Effect |
| --- | --- |
| `includeReference` | also create a dummy column for the reference level |
| `imputeMissing` | `TRUE` sets missing rows to `0` (the reference); `FALSE` (default) keeps `missVal` |
| `dropOriginal` | remove the raw column after encoding |
| `sep` | separator in the new column names (default `"_"`) |

`oneHotEncode()` is idempotent: a dummy column that already exists and agrees with
the requested encoding is left untouched; one that disagrees raises an error. This
makes it safe to run on data that already carries FREM dummy columns.

# 2. The parameter function: `createParamFunction`

The parameter function turns a vector of thetas and one row of covariate values
into the parameters to plot. Writing it by hand duplicates code that already
exists in the control stream, and it is the easiest place in the whole workflow to
introduce a silent error.

## 2.1 What a parameter function must do

Before generating one, it is worth knowing what is being generated — both to
read the output critically and to write one yourself when the generator cannot
(Section 2.3).

The contract is:

```r
function(thetas, df, ...) {
  # ... rebuild the model's typical values ...
  list(CL, FREL, AUC, V)
}
```

* `thetas` is a numeric vector, in the order the THETAs appear in the model.
* `df` is a **one-row** data frame of covariate values — one Forest plot row.
  Because it is one row, `if (df$WT != -99)` is a scalar condition, which is
  what makes the guards below legal.
* `...` absorbs anything else `getForestDFSCM()` passes through, such as the
  extra arguments the [time-to-event vignette](Part3-deep-dive-tte-models.html)
  uses.
* The return value is a list of the quantities to plot, named by
  `functionListName`.

Two defensive checks apply to every covariate the function reads:

* `any(names(df) == "WT")` — the column may be **absent entirely**.
* `df$WT != -99` — the cell may hold the missing-value indicator, which `dfCovs`
  uses for every covariate that is *inactive on that row*.

The second is the one that surprises people. **It is needed even when the
original NONMEM data has no missing covariate values at all**, because `-99` here
does not mean "missing in the data" — it means "this row is not about this
covariate". A `dfCovs` row that varies `WT` carries `-99` in `SEX`, `FOOD` and
the genotype columns, and a function that reads `df$SEX` without the guard will
compute a covariate effect for a row that was never meant to have one.

Written out for a single covariate, the shape is:

```r
FRELSEX <- 1
if (any(names(df) == "SEX") && df$SEX != -99 && df$SEX == 2) {
  FRELSEX <- 1 + thetas[14]
}
```

Note the ordering: the column check comes first, so `df$SEX` is only evaluated
once it is known to exist. Note also `df[["SEX"]]` would be safer still — `$`
partial-matches on a data frame, so a data set carrying `SEXF` but not `SEX`
would silently supply the wrong column. The generated function uses the
bracket form for exactly this reason.

A full hand-written function for `run7`, kept deliberately as a comparison
against the generated one, is in `inst/validation/compare-paramFunction.R` in
the installed package.

## 2.2 Generating one

`createParamFunction()` translates the `$PK` block into R **source text** for you
to read, check and edit. Nothing is evaluated on your behalf.

```{r}
gen <- createParamFunction(
  modFile,
  parameters = c("CL", "FREL", "V"),
  covRef     = list(GENO1 = 0, GENO3 = 0),
  quiet      = TRUE
)
cat(head(gen$code, 24), sep = "\n")
```

The final estimates are what the function will be evaluated with, here and in
`getForestDFSCM()`:

```{r}
ext <- getExt(system.file("extdata", "SimVal/run7.ext", package = "PMXForest"))
thetas <- as.numeric(ext[ext$ITERATION == -1000000000, 2:15])
```

Two things to notice.

**Every `ETA(n)` is set to 0**, so the function returns typical values — what a
Forest plot needs.

**All missing-covariate handling is hoisted into one preamble**, annotated with
where each reference came from, after which the `$PK` algebra below is a plain
transliteration you can diff line by line against the control stream. The
references are read out of the control stream itself, by four rules in decreasing
order of confidence:

| Rule | Source |
| --- | --- |
| 1 | explicit handling, `IF(WT.EQ.-99) WT = 75` |
| 2a | the branch PsN's scm marks `; Most common` |
| 2b | the branch assigning the identity value (`1` multiplicative, `0` additive) |
| 3 | the normalisation constant in `(WT/75)` or `(AGE-50)` |

```{r}
gen$covRef$WT
```

A weaker fifth rule proposes the level that no `IF()` tests and **warns**;
confirm it through `covRef`, as above for `GENO1` and `GENO3`. A covariate
matching no rule at all is an error, not a guess.

## 2.3 When the generator refuses

The parser accepts assignments, `IF` constructs and closed-form arithmetic. It
refuses anything it cannot translate faithfully, naming the file and line rather
than guessing:

| Refused | Why |
| --- | --- |
| `A(1)` — a compartment amount | needs an ODE solution, which no closed-form function can supply |
| `DO`, `WHILE`, `CALL`, `GOTO`, `RETURN` | control flow with no expression equivalent |
| verbatim FORTRAN | not NM-TRAN, so not parseable as such |
| `EPS()` / `ERR()` | residual error has no place in a typical-value function |
| a symbol read before anything assigns it | see below |

That last one is worth explaining, because NONMEM accepts it and R will not.
NONMEM does not initialise `$PK` variables, and it does not clear them between
data records — so a variable read before it is assigned holds whatever the
*previous subject* left there. A parameter function is evaluated one row at a
time with no such carry-over, so it cannot reproduce that behaviour, and
`createParamFunction()` refuses rather than quietly substituting a zero. If you
hit this, give the variable an unconditional value in `$PK` above the line that
reads it.

A branch chain with no `ELSE` is *accepted*, because `scm` writes exhaustive
ones — `run7.mod` itself sets `FRELFORM` from two one-line `IF`s covering
`FORM` 0 and 1 — and no static check can tell an exhaustive chain from a
non-exhaustive one. If no branch fires at run time the generated function stops
with `object not found`, which is the honest outcome.

When the generator refuses, write the function by hand to the contract in
Section 2.1. Two vignettes do exactly that, for models the generator cannot
reach:

* [Time-to-event models](Part3-deep-dive-tte-models.html) — hazard and CDF
  functions, including passing extra arguments through `...`.
* [R-coded models](Part3-deep-dive-r-coded-models.html) — a model that never had
  a `$PK` block to parse.

## 2.4 Adjusting the generated function

The generator returns **source text**, which is the reason it returns text at
all: a function you disagree with can be corrected without abandoning the
translation that was right.

Section 2.3 noted one case where the generated function is faithful to the
control stream and still not enough. `run7`'s `$PK` sets the sex effect with two
one-line branches and no `ELSE`:

```{r}
cat(grep("FRELSEX", gen$code, value = TRUE), sep = "\n")
```

That is exhaustive for the data the model was fitted to, where `SEX` is 1 or 2.
It is not exhaustive in general, and NONMEM's implicit carry-over between records
hides the difference. In R there is nothing to carry over:

```{r, error = TRUE}
paramFunction <- eval(parse(text = paste(gen$code, collapse = "\n")))
paramFunction(thetas, df = data.frame(SEX = 3))
```

The repair is one line, inserted where the branches begin — and inserted by
matching on the text rather than by counting lines, so it still lands in the
right place after the model is re-run and the source regenerated:

```{r}
code <- gen$code
at <- grep("if (SEX == 1) FRELSEX <- 1", code, fixed = TRUE)
code <- append(code, "  FRELSEX <- 1  # default the scm branches leave implicit", after = at - 1)

paramFunctionFixed <- eval(parse(text = paste(code, collapse = "\n")))
```

The unmodelled level now resolves to the reference, and the modelled ones are
untouched — which is the property to check after any edit of this kind:

```{r}
vapply(1:3, function(s) paramFunctionFixed(thetas, data.frame(SEX = s))$FREL, numeric(1))
```

Three habits make this safe rather than a fork of the model:

**Edit the object, not a copy.** `gen$code` is a character vector, so the change
is `sub()`, `grep()` and `append()` in the script. Pasting the generated source
into the file and hand-editing it there produces something that looks
hand-written, cannot be regenerated, and will silently disagree with the model
after the next run.

**Match on content, not position.** `code[24]` is correct until a covariate is
added to `$PK`. `grep(..., fixed = TRUE)` on the line's text still finds it, and
`length(at) == 1` is worth asserting if the pattern could match twice.

**Say why, in the code.** The inserted line carries a comment. When someone asks
in six months why the Forest plot does not match the control stream exactly, the
answer is in the script rather than in anyone's memory.

If the edit grows past a few lines, that is a signal the model needs the
correction rather than the plot — fix `$PK` and regenerate.

## Secondary parameters are yours

AUC, Cmax and event probabilities are not derivable from `$PK` — the dose used
for an AUC appears nowhere in the control stream, and anything through `$ERROR`
or `$DES` needs an ODE solution. The emitted source carries a marked extension
point for you to add them:

```{r}
cat(grep("Secondary parameters", gen$code, value = TRUE))
```

You do not have to add them by editing the emitted text. `secondary` takes them
as an argument, either as a line of R code or as the path to a whole `.R` file —
an `mrgsolve` or `deSolve` simulation, say — which is inlined so the result
stays self-contained. That is a topic in itself, and it has its own vignette:
[Secondary parameters](Part3-deep-dive-secondary-parameters.html).

## Checking the translation: `verifyParamFunction()`

`verifyParamFunction()` evaluates the generated function over a NONMEM `$TABLE`
file and compares the result with the values NONMEM itself wrote, which turns
"do I trust this translation?" into a pass or a fail.

```r
v <- verifyParamFunction(gen, tabFile = "xptab7", thetas = thetas)
v
#> PASS - verifyParamFunction: 2/2 parameter(s)
#>  PARAMETER      TABLECOLUMN   N   MAXABSDIFF   MAXRELDIFF PASS
#>         CL   CL / exp(ETA3) 748 0.0006296369 4.848771e-05 TRUE
#>       FREL FREL / exp(ETA2) 748 0.0001012091 5.217360e-05 TRUE
```

The value is a single `TRUE`/`FALSE` - `TRUE` only if every requested
parameter was checked and passed - so it drops straight into
`if (verifyParamFunction(...)) ...`. The per-parameter table shown above is
attached as `attr(v, "checks")`.

Put **both the parameters and the covariates** in `$TABLE`. They then come from
the same row of the same file and cannot be misaligned, and no `$DATA` filtering
has to be reproduced to compare them. Where `$PK` shows exponential IIV, the
tabled individual value is divided by `exp(ETA(n))` to recover the typical value;
a `TVCL`-style column is used directly. Anything missing warns and reports `NA`
rather than comparing silently.

::: {.callout-note}
The residual above is not error, it is the table's precision: NONMEM's default
format carries about five significant digits, so agreement closer than roughly
`1e-5` relative cannot be observed. `tol` defaults to `1e-4`, just above that
floor.
:::

# 3. The reference row: `dfRefRow`

`setupDfRefRow()` builds the reference the other rows are expressed relative to.
By default it computes the reference state from the data — mode for categorical,
median or mean for continuous, on one record per subject — and maps it onto the
geometry of a `setupDfCovs()` result.

`singleRef = TRUE` gives a one-row data frame with every covariate at its
reference value:

```{r}
setupDfRefRow(dfCovs, data = dfData, covariates = covariates, idVar = "ID")
```

`singleRef = FALSE` gives a data frame the same shape as `dfCovs`: active cells
are replaced with the reference value, inactive cells keep `-99`. Use it when the
reference should track the covariate that varies on each row.

```{r}
head(setupDfRefRow(dfCovs,
  data = dfData, covariates = covariates,
  singleRef = FALSE, idVar = "ID"
), 4)
```

## `contRef` and `catRef`

These say how a reference value is chosen. Each takes either a single setting
applied to every covariate, or a named list with an entry per covariate and an
optional `default` component:

| | accepted values |
| --- | --- |
| `contRef` | a number, `"mean"`, `"median"` (default), `"model"` |
| `catRef` | a level, `"mode"` (default), `"lowest"`, `"model"` |

```{r}
setupDfRefRow(dfCovs, dfData, covariates,
  idVar = "ID",
  contRef = list(WT = 75, default = "mean")
)
```

`contRef` applies to `setupDfCovs()` too — to `additionalCovs`, and to primary
covariates when `useMissVal = FALSE`.

### `"model"` needs the covariate to be in `$PK`

`"model"` reads the reference out of the control stream, by the same rules
Section 2 lists. That makes it the setting that keeps the reference row and the
parameter function consistent — Section 4 is about why that matters — but it
only works for covariates `$PK` actually mentions.

**As a bare scalar it applies to every covariate, so one unmentioned covariate
fails the whole call.** `run7`'s `$PK` never mentions `AGE`:

```r
setupDfRefRow(dfCovs, dfData, covariates, idVar = "ID", contRef = "model", model = modFile)
#> Error: No reference value for covariate AGE could be derived from the model.
#> Give it explicitly, e.g. contRef = list(AGE = <value>).
```

Use the named-list form and let everything else fall back:

```r
contRef = list(WT = "model", default = "median")
```

**A one-hot coded categorical needs a literal level, not `"model"`.** `$PK`
names the dummies `GENO1`, `GENO3`, `GENO4` — never bare `GENO` — so the model
holds a reference for each dummy but none for the covariate they encode:

```r
catRef = list(GENO = "model")
#> Error: No reference value for covariate GENO could be derived from the model.
#> Give it explicitly, e.g. catRef = list(GENO = <level>).
#> The $PK names GENO4, GENO1, GENO3 rather than GENO itself, so a reference was
#> found for each of those but not for the covariate they encode. Give the level
#> you want as the reference.
```

So `catRef = list(SEX = "model", FORM = "model", FOOD = "model", GENO = 2)`:
`SEX`, `FORM` and `FOOD` are coded directly in `$PK` and resolve; `GENO` is
given the level you want.

::: {.callout-warning}
`setupDfRefRow()` keys the reference values by the column names `setupDfCovs()`
produces. Pass it a `dfCovs` built by `setupDfCovs()`, and the **same** `catRef`
and `sep`, so the columns line up. For a hand-built `createInputForestData()`
structure, construct `dfRefRow` directly as a one-row data frame.
:::

# 4. Keeping the reference and the parameter function consistent

This section is the reason the previous two sit in the same vignette.

A parameter function must return something when a covariate is inactive, and it
falls back to the model's own reference — the `75` in `(WT/75)`, or the level in
the branch marked `; Most common`. If `setupDfRefRow()` meanwhile takes the median
weight and the modal formulation from the data, the two disagree. Every row where
that covariate is inactive is then displaced from 1, by a factor that looks
entirely plausible on the plot.

For `run7` the two disagree on both counts:

```{r}
covsMod <- c("WT", "SEX", "FOOD", "FORM")
dfCovsM <- setupDfCovs(dfData, covariates = covsMod, idVar = "ID")

## From the data
setupDfRefRow(dfCovsM, dfData, covsMod, idVar = "ID")

## From the control stream
setupDfRefRow(dfCovsM, dfData, covsMod,
  idVar = "ID",
  contRef = "model", catRef = "model", model = modFile
)
```

`WT` is 85.4 in the data and 75 in the model; `FORM` is 0 in the data and 1 in the
model. With the data-derived reference, every row that does not vary `WT` sits at
`(75/85.4)^theta` instead of 1 — 0.907 for CL and 0.878 for V — and every row that
does not vary `FORM` sits at 1.17 for Frel:

```{r}
c(CL = (75 / 85.4)^thetas[2], V = (75 / 85.4)^thetas[3], Frel = 1 / (1 + thetas[12]))
```

Setting `contRef = "model"` and `catRef = "model"` removes the mismatch by
construction. `model` takes either a control stream path or the list returned by
`createParamFunction()`; passing the latter guarantees the reference row and the
parameter function come from one derivation.

```{r}
dfRefRow <- setupDfRefRow(dfCovsM, dfData, covsMod,
  idVar = "ID",
  contRef = "model", catRef = "model", model = gen
)
dfRefRow
```

::: {.callout-note}
`inst/validation/compare-paramFunction.R` in the installed package draws the two
Forest plots side by side — a hand-written parameter function and a generated one
— and reports the largest difference between them. It is a useful template for
checking a translation on your own model.
:::

# 5. Which records to summarise

Everything above summarises whatever data frame you hand it. NONMEM, meanwhile,
reads only the records its `$DATA` `IGNORE` and `ACCEPT` statements leave, so the
analysis file and the model's population are not the same thing.

```{r}
dfUsed <- filterByModel(dfData, modFile, quiet = TRUE)

c(
  records_file = nrow(dfData), records_model = nrow(dfUsed),
  subjects_file = length(unique(dfData$ID)),
  subjects_model = length(unique(dfUsed$ID))
)
```

Over a fifth of the subjects in the file were never seen by `run7`, and that moves
the quantiles the plot is built from:

```{r}
getCovStats(dfData, "CRCL", idVar = "ID")
getCovStats(dfUsed, "CRCL", idVar = "ID")
```

So summarise `filterByModel(dfData, modFile)` rather than the raw file — or pass a
`$TABLE` output, which is already filtered.

::: {.callout-warning}
`filterByModel()` matches columns **by position, not by name**. NONMEM skips the
header line and takes its names from `$INPUT`, so the two can disagree — and in
this very data set they do:

```{r}
mod <- PMXForest:::nmReadModel(modFile)
data.frame(
  position = 9:10,
  input = PMXForest:::nmInputPositions(mod)$names[9:10],
  csv_header = names(dfData)[9:10]
)
```

`$INPUT`'s `DV` is the file's `LNDV`. Filtering by name would read the wrong
column. `useInputNames = TRUE` returns the data as NONMEM sees it, renamed.
:::

The single-character form (`IGNORE=C`) is a rule about the first character of each
raw record rather than a condition on the data, so it cannot be applied to a data
frame; it is skipped with a warning. The conventional `@` and `#` header markers
are skipped silently, since `read.csv()` has already dealt with them.

# 6. Parameter uncertainty: `dfParameters`

`getSamples()` builds the sampled parameter vectors, from a `.cov` file (a
multivariate normal draw around the estimates), a PsN bootstrap, a SIR
`raw_results` file, or a data frame you supply.

```{r}
dfSamples <- getSamples(
  system.file("extdata", "SimVal/run7.cov", package = "PMXForest"),
  system.file("extdata", "SimVal/run7.ext", package = "PMXForest"),
  n = 25
)
dim(dfSamples)
```

The first row is always the final estimates, so row 1 of the Forest plot data is
the point estimate and the remaining rows give the interval. `noBaseThetas` says
how many leading columns to pass to the parameter function; `createParamFunction()`
reports the right value for the model it read.

```{r}
gen$noBaseThetas
```

See the [Walkthrough](Part2-walkthrough.html) for the full treatment, including
bootstrap and SIR inputs.

# 7. Empirical Forest plots: `covExpressionsList`

`getForestDFemp()` takes a named list of `expression()` objects, one per Forest
plot row, each selecting the subjects it summarises.
`setupCovExpressionsList()` builds that list — and the matching `cdfCovsNames`
labels — from the same deduplicated statistics `setupDfCovs()` uses.

```{r}
emp <- setupCovExpressionsList(
  data       = dfUsed,
  covariates = c("WT", "SEX", "GENO"),
  idVar      = "ID"
)
emp$covExpressionsList
emp$cdfCovsNames
```

Continuous covariates become two rows at the `probs` quantile tails
(`WT < q_05`, `WT >= q_95`); `contSplit = "median"` splits at the median instead
so every subject falls in one row. Binary and multi-level categorical covariates
become one `== level` row per level; `includeReference = FALSE` drops the
reference level, as in the parametric `dfCovs`.

An `additionalCovs` entry gets its own rows *and* contributes a fixed condition
that is combined with `&` into every other covariate's expression: a level for a
categorical covariate, or a `prob` / `value` split with a direction for a
continuous one. Use it to restrict the empirical contrast to a sub-population.

```{r}
setupCovExpressionsList(
  data           = dfUsed,
  covariates     = c("WT", "SEX"),
  contSplit      = "median",
  additionalCovs = list(FOOD = 1, CRCL = list(prob = 0.25, dir = "gt")),
  idVar          = "ID"
)$covExpressionsList
```

Every generated expression is evaluated against the deduplicated data. If any
selects fewer than `minSubjects` subjects (default 10) the function stops, so an
empty or near-empty subset never reaches `getForestDFemp()`.

```{r, error = TRUE}
setupCovExpressionsList(dfUsed, "WT", probs = c(0.001, 0.999), idVar = "ID")
```

The reference for an empirical plot stays deterministic: build it with
`setupDfCovs()` and `setupDfRefRow(singleRef = TRUE)` and pass the one-row data
frame to `getForestDFemp(dfRefRow = )`, exactly as for a parametric plot.

# 8. Missing covariate values

`getCovStats()` and the setup functions exclude `missVal` before computing
statistics, so a covariate that is missing for some subjects is summarised from
the subjects who have it.

```{r}
dfMiss <- data.frame(
  ID  = 1:8,
  WT  = c(60.5, 70.2, 70.5, 80.8, 65.1, 90.0, 77.4, 68.9),
  SEX = c(0, 1, 1, 0, 1, 0, 1, 0),
  BMI = c(22.1, 24.4, -99, 27.3, 23.5, 28.9, -99, 25.1)
)

getCovStats(dfMiss, c("WT", "SEX", "BMI"), idVar = "ID", minLevels = 4)
```

The `BMI` quantiles are computed from the six non-missing values. The parameter
function still has to handle `-99` in the `BMI` column of `dfCovs`, because that
is what marks the rows where `BMI` is inactive — which is exactly what the
preamble in Section 2 does.

# 9. Feeding the result forward

Putting the four inputs together:

```r
dfUsed    <- filterByModel(dfData, modFile)
dfCovs    <- setupDfCovs(dfUsed, covariates = covs, idVar = "ID")
gen       <- createParamFunction(modFile, parameters = c("CL", "V"))
paramFunc <- eval(parse(text = gen$code))          # after reading it
dfRefRow  <- setupDfRefRow(dfCovs, dfUsed, covs, idVar = "ID",
                           contRef = "model", catRef = "model", model = gen)
dfSamples <- getSamples(covFile, extFile, n = 1000)

dfres <- getForestDFSCM(
  dfCovs           = dfCovs,
  functionList     = list(paramFunc),
  functionListName = gen$functionListName,
  noBaseThetas     = gen$noBaseThetas,
  dfParameters     = dfSamples,
  dfRefRow         = dfRefRow
)
```

See the [Walkthrough](Part2-walkthrough.html) for the full call and the plotting
step.

For empirical Forest plots, `getForestDFemp()` takes `covExpressionsList` in place
of `dfCovs`. If the parameter function is written against one-hot dummy columns
(`GENO_2`, `GENO_3`, ...) rather than the raw `GENO`, give `getForestDFemp()` its
own `oneHot` argument — the same spec `oneHotEncode()` takes — so those columns
are added to the data passed in before the parameter function runs. The expressions
from `setupCovExpressionsList()` always filter on the raw column, so that choice
is independent of how the expression list is built.
