---
title: "3. Deep Dive: Preparing Forest Plot Inputs"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Preparing Forest Plot Inputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
.callout-warning {
  background-color: #fff3cd;
  border-left: 4px solid #ffc107;
  color: #664d03;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(dplyr)
})

set.seed(865765)
```

# Introduction

`getForestDFSCM()` takes four inputs. This vignette covers how each is built, and
— just as important — how to keep them consistent with one another.

| Input | What it is | Built by |
| --- | --- | --- |
| `dfCovs` | one row per Forest plot row, inactive cells at `-99` | `setupDfCovs()`, `createInputForestData()` |
| `functionList` | the parameter function, typical values from thetas and covariates | `createParamFunction()`, or written by hand |
| `dfRefRow` | the reference subject the other rows are expressed relative to | `setupDfRefRow()` |
| `dfParameters` | parameter uncertainty samples | `getSamples()` |

The empirical entry point `getForestDFemp()` takes a `covExpressionsList` in place
of `dfCovs`; that is covered in Section 7.

The data set throughout is the PK analysis data used in the
[Walkthrough](Part2-walkthrough.html), and the model is the `run7` control stream
bundled with the package.

```{r}
dataFile <- system.file("extdata", "SimVal/DAT-1-MI-PMX-2.csv", package = "PMXForest")
modFile  <- system.file("extdata", "SimVal/run7.mod",           package = "PMXForest")
dfData   <- read.csv(dataFile)
```

# 1. Covariate values: `dfCovs`

## `getCovStats()`

`getCovStats()` returns a named list, one element per covariate. The element type
depends on the number of unique values:

* **continuous** (more than `minLevels` unique values) — a two-element numeric
  vector of quantiles, by default the 5th and 95th percentiles;
* **binary** — the two levels, sorted alphanumerically;
* **multi-level categorical** — a nested list of one-hot vectors. The reference
  level (the lowest level, or the one named in `catRef`) is the all-zero row.

```{r}
covStats <- getCovStats(
  data       = dfData,
  covariates = c("WT", "AGE", "SEX", "GENO"),
  idVar      = "ID"
)
covStats
```

Statistics are computed on the first record per `idVar`, so subjects with dense
sampling do not pull the quantiles.

```{r}
nrow(dfData)                                   # all records
nrow(distinct(dfData, ID))                     # subjects
```

`minLevels` (default 10) is the threshold between categorical and continuous. A
covariate coded with a small number of distinct numeric values — a dose level, an
ordinal score — is treated as categorical unless `minLevels` is lowered.

```{r}
## GENO has 4 levels: categorical by default
getCovStats(dfData, "GENO", idVar = "ID")

## Force GENO to be treated as continuous
getCovStats(dfData, "GENO", idVar = "ID", minLevels = 3)
```

`probs` sets the quantiles for continuous covariates; `nsig` sets the rounding.

```{r}
getCovStats(dfData, "CRCL", idVar = "ID", probs = c(0.1, 0.9), nsig = 4)
```

## `catRef` and `sep`

By default the lowest level of a multi-level categorical is the reference (the
all-zero row) and the one-hot columns are named `<covariate>_<level>`.
`catRef` overrides the reference level per covariate; `sep` changes the
separator. (`catRef` replaces `refLevels`, which still works but is deprecated.)

```{r}
## GENO reference is level 1 by default -> columns GENO_2, GENO_3, GENO_4
getCovStats(dfData, "GENO", idVar = "ID")

## Make level 2 the reference -> columns GENO_1, GENO_3, GENO_4
getCovStats(dfData, "GENO", idVar = "ID", catRef = list(GENO = 2))
```

Use `catRef` when the model's reference category is not the lowest level —
otherwise the one-hot columns will not line up with the model's coefficients. The
`"_"` separator matches NONMEM FREM data sets and `PMXFrem::createFREMModel()`;
change it only if you have a reason to.

::: {.callout-note}
Binary levels are returned in alphanumeric order (`c(0, 1)`), not order of
appearance in the data. Forest plot row order is therefore reproducible across
data sets.
:::

## `createInputForestData()`

`createInputForestData()` turns a covariate list into the row structure. It is
what you call directly when the covariate values are known literals rather than
derived from data.

```{r}
dfCovsManual <- createInputForestData(
  list(
    "WT"   = c(60, 100),
    "SEX"  = c(1, 2),
    "GENO" = list("GENO_2" = c(0, 1, 0), "GENO_3" = c(0, 0, 1), "GENO_4" = c(0, 0, 0))
  ),
  iMiss = -99
)
dfCovsManual
```

The `COVARIATEGROUPS` column records which covariate each row belongs to;
`getForestDFSCM()` uses it to group and space the rows on the y-axis, and drops it
before calling the parameter function.

A `getCovStats()` result can be passed straight in:

```{r}
createInputForestData(covStats, iMiss = -99)
```

## `setupDfCovs()`

`setupDfCovs()` runs `getCovStats()` and `createInputForestData()` together and
adds control over how inactive cells are filled.

```{r}
covariates <- c("WT", "AGE", "CRCL", "SEX", "NCIL", "GENO")

dfCovs <- setupDfCovs(data = dfData, covariates = covariates, idVar = "ID")
dfCovs
```

Covariates listed in `additionalCovs` get their own rows, but on every *other*
row they sit at their reference rather than at `-99`. This is for covariates that
must hold a plausible value whenever they are not the covariate under study — for
example one carried through from a full random-effects (FREM) model.

```{r}
setupDfCovs(
  data           = dfData,
  covariates     = c("WT", "AGE", "SEX"),
  additionalCovs = "CRCL",
  idVar          = "ID"
)
```

On the `WT`, `AGE`, and `SEX` rows, `CRCL` holds its median instead of `-99`.

With `useMissVal = FALSE`, inactive *primary* covariates are also set to their
reference, so the parameter function never sees `-99`.

```{r}
setupDfCovs(
  data       = dfData,
  covariates = c("WT", "AGE", "SEX"),
  useMissVal = FALSE,
  idVar      = "ID"
)
```

## `oneHotEncode()`

`getCovStats()` produces one-hot vectors from a *summary* of the data.
`oneHotEncode()` does the same job on a *data frame*: it adds
`<covariate><sep><level>` dummy columns to whatever data frame you give it. This
is what the `oneHot` argument of `getForestDFSCM()` and `getForestDFemp()` calls
internally, and it follows the same convention as NONMEM FREM data sets.

```{r}
raw <- data.frame(
  ID   = 1:6,
  GENO = c(1, 2, 3, 4, 2, -99),
  RACE = c(1, 1, 2, 3, 2, 1)
)

## Lowest level as reference; missing rows kept as missVal
oneHotEncode(raw, spec = c("GENO", "RACE"))
```

The reference level is set the same way as in `getCovStats()`:

```{r}
oneHotEncode(raw, spec = list(GENO = list(ref = 2)), dropOriginal = TRUE)
```

| Argument | Effect |
| --- | --- |
| `includeReference` | also create a dummy column for the reference level |
| `imputeMissing` | `TRUE` sets missing rows to `0` (the reference); `FALSE` (default) keeps `missVal` |
| `dropOriginal` | remove the raw column after encoding |
| `sep` | separator in the new column names (default `"_"`) |

`oneHotEncode()` is idempotent: a dummy column that already exists and agrees with
the requested encoding is left untouched; one that disagrees raises an error. This
makes it safe to run on data that already carries FREM dummy columns.

# 2. The parameter function: `createParamFunction`

The parameter function turns a vector of thetas and one row of covariate values
into the parameters to plot. Writing it by hand duplicates code that already
exists in the control stream, and it is the easiest place in the whole workflow to
introduce a silent error.

`createParamFunction()` translates the `$PK` block into R **source text** for you
to read, check and edit. Nothing is evaluated on your behalf.

```{r}
gen <- createParamFunction(
  modFile,
  parameters = c("CL", "FREL", "V"),
  covRef     = list(GENO1 = 0, GENO3 = 0),
  quiet      = TRUE
)
cat(head(gen$code, 24), sep = "\n")
```

Two things to notice.

**Every `ETA(n)` is set to 0**, so the function returns typical values — what a
Forest plot needs.

**All missing-covariate handling is hoisted into one preamble**, annotated with
where each reference came from, after which the `$PK` algebra below is a plain
transliteration you can diff line by line against the control stream. The
references are read out of the control stream itself, by four rules in decreasing
order of confidence:

| Rule | Source |
| --- | --- |
| 1 | explicit handling, `IF(WT.EQ.-99) WT = 75` |
| 2a | the branch PsN's scm marks `; Most common` |
| 2b | the branch assigning the identity value (`1` multiplicative, `0` additive) |
| 3 | the normalisation constant in `(WT/75)` or `(AGE-50)` |

```{r}
gen$covRef$WT
```

A weaker fifth rule proposes the level that no `IF()` tests and **warns**;
confirm it through `covRef`, as above for `GENO1` and `GENO3`. A covariate
matching no rule at all is an error, not a guess.

## Secondary parameters are yours

AUC, Cmax and event probabilities are not derivable from `$PK` — the dose used
for an AUC appears nowhere in the control stream, and anything through `$ERROR`
or `$DES` needs an ODE solution. The emitted source carries a marked extension
point for you to add them:

```{r}
cat(grep("Secondary parameters", gen$code, value = TRUE))
```

The parser accepts assignments, `IF` statements and closed-form arithmetic, and
refuses everything else — `$DES`, compartment amounts `A(n)`, verbatim FORTRAN,
`DO` loops, `CALL` — naming the file and line rather than translating it wrongly.

## Checking the translation: `verifyParamFunction()`

`verifyParamFunction()` evaluates the generated function over a NONMEM `$TABLE`
file and compares the result with the values NONMEM itself wrote, which turns
"do I trust this translation?" into a pass or a fail.

```r
rep <- verifyParamFunction(gen, tabFile = "xptab7", thetas = thetas)
#>   PARAMETER      TABLECOLUMN   N   MAXABSDIFF   MAXRELDIFF PASS
#> 1        CL   CL / exp(ETA3) 748 0.0006296369 4.848771e-05 TRUE
#> 2      FREL FREL / exp(ETA2) 748 0.0001012091 5.217360e-05 TRUE
```

Put **both the parameters and the covariates** in `$TABLE`. They then come from
the same row of the same file and cannot be misaligned, and no `$DATA` filtering
has to be reproduced to compare them. Where `$PK` shows exponential IIV, the
tabled individual value is divided by `exp(ETA(n))` to recover the typical value;
a `TVCL`-style column is used directly. Anything missing warns and reports `NA`
rather than comparing silently.

::: {.callout-note}
The residual above is not error, it is the table's precision: NONMEM's default
format carries about five significant digits, so agreement closer than roughly
`1e-5` relative cannot be observed. `tol` defaults to `1e-4`, just above that
floor.
:::

# 3. The reference row: `dfRefRow`

`setupDfRefRow()` builds the reference the other rows are expressed relative to.
By default it computes the reference state from the data — mode for categorical,
median or mean for continuous, on one record per subject — and maps it onto the
geometry of a `setupDfCovs()` result.

`singleRef = TRUE` gives a one-row data frame with every covariate at its
reference value:

```{r}
setupDfRefRow(dfCovs, data = dfData, covariates = covariates, idVar = "ID")
```

`singleRef = FALSE` gives a data frame the same shape as `dfCovs`: active cells
are replaced with the reference value, inactive cells keep `-99`. Use it when the
reference should track the covariate that varies on each row.

```{r}
head(setupDfRefRow(dfCovs, data = dfData, covariates = covariates,
                   singleRef = FALSE, idVar = "ID"), 4)
```

## `contRef` and `catRef`

These say how a reference value is chosen. Each takes either a single setting
applied to every covariate, or a named list with an entry per covariate and an
optional `default` component:

| | accepted values |
| --- | --- |
| `contRef` | a number, `"mean"`, `"median"` (default), `"model"` |
| `catRef` | a level, `"mode"` (default), `"lowest"`, `"model"` |

```{r}
setupDfRefRow(dfCovs, dfData, covariates, idVar = "ID",
              contRef = list(WT = 75, default = "mean"))
```

`contRef` applies to `setupDfCovs()` too — to `additionalCovs`, and to primary
covariates when `useMissVal = FALSE`.

::: {.callout-warning}
`setupDfRefRow()` keys the reference values by the column names `setupDfCovs()`
produces. Pass it a `dfCovs` built by `setupDfCovs()`, and the **same** `catRef`
and `sep`, so the columns line up. For a hand-built `createInputForestData()`
structure, construct `dfRefRow` directly as a one-row data frame.
:::

# 4. Keeping the reference and the parameter function consistent

This section is the reason the previous two sit in the same vignette.

A parameter function must return something when a covariate is inactive, and it
falls back to the model's own reference — the `75` in `(WT/75)`, or the level in
the branch marked `; Most common`. If `setupDfRefRow()` meanwhile takes the median
weight and the modal formulation from the data, the two disagree. Every row where
that covariate is inactive is then displaced from 1, by a factor that looks
entirely plausible on the plot.

For `run7` the two disagree on both counts:

```{r}
covsMod <- c("WT", "SEX", "FOOD", "FORM")
dfCovsM <- setupDfCovs(dfData, covariates = covsMod, idVar = "ID")

## From the data
setupDfRefRow(dfCovsM, dfData, covsMod, idVar = "ID")

## From the control stream
setupDfRefRow(dfCovsM, dfData, covsMod, idVar = "ID",
              contRef = "model", catRef = "model", model = modFile)
```

`WT` is 85.4 in the data and 75 in the model; `FORM` is 0 in the data and 1 in the
model. With the data-derived reference, every row that does not vary `WT` sits at
`(75/85.4)^theta` instead of 1 — 0.907 for CL and 0.878 for V — and every row that
does not vary `FORM` sits at 1.17 for Frel:

```{r}
ext    <- getExt(system.file("extdata", "SimVal/run7.ext", package = "PMXForest"))
thetas <- as.numeric(ext[ext$ITERATION == -1000000000, 2:15])

c(CL = (75 / 85.4)^thetas[2], V = (75 / 85.4)^thetas[3], Frel = 1 / (1 + thetas[12]))
```

Setting `contRef = "model"` and `catRef = "model"` removes the mismatch by
construction. `model` takes either a control stream path or the list returned by
`createParamFunction()`; passing the latter guarantees the reference row and the
parameter function come from one derivation.

```{r}
dfRefRow <- setupDfRefRow(dfCovsM, dfData, covsMod, idVar = "ID",
                          contRef = "model", catRef = "model", model = gen)
dfRefRow
```

::: {.callout-note}
`inst/validation/compare-paramFunction.R` in the installed package draws the two
Forest plots side by side — a hand-written parameter function and a generated one
— and reports the largest difference between them. It is a useful template for
checking a translation on your own model.
:::

# 5. Which records to summarise

Everything above summarises whatever data frame you hand it. NONMEM, meanwhile,
reads only the records its `$DATA` `IGNORE` and `ACCEPT` statements leave, so the
analysis file and the model's population are not the same thing.

```{r}
dfUsed <- filterByModel(dfData, modFile, quiet = TRUE)

c(records_file = nrow(dfData),  records_model = nrow(dfUsed),
  subjects_file = length(unique(dfData$ID)),
  subjects_model = length(unique(dfUsed$ID)))
```

Over a fifth of the subjects in the file were never seen by `run7`, and that moves
the quantiles the plot is built from:

```{r}
getCovStats(dfData, "CRCL", idVar = "ID")
getCovStats(dfUsed, "CRCL", idVar = "ID")
```

So summarise `filterByModel(dfData, modFile)` rather than the raw file — or pass a
`$TABLE` output, which is already filtered.

::: {.callout-warning}
`filterByModel()` matches columns **by position, not by name**. NONMEM skips the
header line and takes its names from `$INPUT`, so the two can disagree — and in
this very data set they do:

```{r}
mod <- PMXForest:::nmReadModel(modFile)
data.frame(position   = 9:10,
           input      = PMXForest:::nmInputPositions(mod)$names[9:10],
           csv_header = names(dfData)[9:10])
```

`$INPUT`'s `DV` is the file's `LNDV`. Filtering by name would read the wrong
column. `useInputNames = TRUE` returns the data as NONMEM sees it, renamed.
:::

The single-character form (`IGNORE=C`) is a rule about the first character of each
raw record rather than a condition on the data, so it cannot be applied to a data
frame; it is skipped with a warning. The conventional `@` and `#` header markers
are skipped silently, since `read.csv()` has already dealt with them.

# 6. Parameter uncertainty: `dfParameters`

`getSamples()` builds the sampled parameter vectors, from a `.cov` file (a
multivariate normal draw around the estimates), a PsN bootstrap, a SIR
`raw_results` file, or a data frame you supply.

```{r}
dfSamples <- getSamples(
  system.file("extdata", "SimVal/run7.cov", package = "PMXForest"),
  system.file("extdata", "SimVal/run7.ext", package = "PMXForest"),
  n = 25
)
dim(dfSamples)
```

The first row is always the final estimates, so row 1 of the Forest plot data is
the point estimate and the remaining rows give the interval. `noBaseThetas` says
how many leading columns to pass to the parameter function; `createParamFunction()`
reports the right value for the model it read.

```{r}
gen$noBaseThetas
```

See the [Walkthrough](Part2-walkthrough.html) for the full treatment, including
bootstrap and SIR inputs.

# 7. Empirical Forest plots: `covExpressionsList`

`getForestDFemp()` takes a named list of `expression()` objects, one per Forest
plot row, each selecting the subjects it summarises.
`setupCovExpressionsList()` builds that list — and the matching `cdfCovsNames`
labels — from the same deduplicated statistics `setupDfCovs()` uses.

```{r}
emp <- setupCovExpressionsList(
  data       = dfUsed,
  covariates = c("WT", "SEX", "GENO"),
  idVar      = "ID"
)
emp$covExpressionsList
emp$cdfCovsNames
```

Continuous covariates become two rows at the `probs` quantile tails
(`WT < q_05`, `WT >= q_95`); `contSplit = "median"` splits at the median instead
so every subject falls in one row. Binary and multi-level categorical covariates
become one `== level` row per level; `includeReference = FALSE` drops the
reference level, as in the parametric `dfCovs`.

An `additionalCovs` entry gets its own rows *and* contributes a fixed condition
that is combined with `&` into every other covariate's expression: a level for a
categorical covariate, or a `prob` / `value` split with a direction for a
continuous one. Use it to restrict the empirical contrast to a sub-population.

```{r}
setupCovExpressionsList(
  data           = dfUsed,
  covariates     = c("WT", "SEX"),
  contSplit      = "median",
  additionalCovs = list(FOOD = 1, CRCL = list(prob = 0.25, dir = "gt")),
  idVar          = "ID"
)$covExpressionsList
```

Every generated expression is evaluated against the deduplicated data. If any
selects fewer than `minSubjects` subjects (default 10) the function stops, so an
empty or near-empty subset never reaches `getForestDFemp()`.

```{r, error = TRUE}
setupCovExpressionsList(dfUsed, "WT", probs = c(0.001, 0.999), idVar = "ID")
```

The reference for an empirical plot stays deterministic: build it with
`setupDfCovs()` and `setupDfRefRow(singleRef = TRUE)` and pass the one-row data
frame to `getForestDFemp(dfRefRow = )`, exactly as for a parametric plot.

# 8. Missing covariate values

`getCovStats()` and the setup functions exclude `missVal` before computing
statistics, so a covariate that is missing for some subjects is summarised from
the subjects who have it.

```{r}
dfMiss <- data.frame(
  ID  = 1:8,
  WT  = c(60.5, 70.2, 70.5, 80.8, 65.1, 90.0, 77.4, 68.9),
  SEX = c(0, 1, 1, 0, 1, 0, 1, 0),
  BMI = c(22.1, 24.4, -99, 27.3, 23.5, 28.9, -99, 25.1)
)

getCovStats(dfMiss, c("WT", "SEX", "BMI"), idVar = "ID", minLevels = 4)
```

The `BMI` quantiles are computed from the six non-missing values. The parameter
function still has to handle `-99` in the `BMI` column of `dfCovs`, because that
is what marks the rows where `BMI` is inactive — which is exactly what the
preamble in Section 2 does.

# 9. Feeding the result forward

Putting the four inputs together:

```r
dfUsed    <- filterByModel(dfData, modFile)
dfCovs    <- setupDfCovs(dfUsed, covariates = covs, idVar = "ID")
gen       <- createParamFunction(modFile, parameters = c("CL", "V"))
paramFunc <- eval(parse(text = gen$code))          # after reading it
dfRefRow  <- setupDfRefRow(dfCovs, dfUsed, covs, idVar = "ID",
                           contRef = "model", catRef = "model", model = gen)
dfSamples <- getSamples(covFile, extFile, n = 1000)

dfres <- getForestDFSCM(
  dfCovs           = dfCovs,
  functionList     = list(paramFunc),
  functionListName = gen$functionListName,
  noBaseThetas     = gen$noBaseThetas,
  dfParameters     = dfSamples,
  dfRefRow         = dfRefRow
)
```

See the [Walkthrough](Part2-walkthrough.html) for the full call and the plotting
step.

For empirical Forest plots, `getForestDFemp()` takes `covExpressionsList` in place
of `dfCovs`. If the parameter function is written against one-hot dummy columns
(`GENO_2`, `GENO_3`, ...) rather than the raw `GENO`, give `getForestDFemp()` its
own `oneHot` argument — the same spec `oneHotEncode()` takes — so those columns
are added to the analysis data before the parameter function runs. The expressions
from `setupCovExpressionsList()` always filter on the raw column, so that choice
is independent of how the expression list is built.
