---
title: "3. Deep Dive: Covariate Data Preparation"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Covariate Data Preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
.callout-warning {
  background-color: #fff3cd;
  border-left: 4px solid #ffc107;
  color: #664d03;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(dplyr)
})

set.seed(865765)
```

# Introduction

The covariate structure passed to `getForestDFSCM()` is a data frame in which each
row is one row of the Forest plot, with inactive cells set to a missing-value
indicator (`-99` by default). Four functions build that structure:

| Function | Role |
| --- | --- |
| `getCovStats()` | summarise each covariate from the data |
| `createInputForestData()` | reshape a covariate list into the row structure |
| `setupDfCovs()` | `getCovStats()` + `createInputForestData()` in one call |
| `setupDfRefRow()` | build the reference row(s) matched to a `setupDfCovs()` result |

This vignette covers each in turn, then the handling of missing covariate values.
The data set is the PK analysis data used in the
[Walkthrough](Part2-walkthrough.html).

```{r}
dataFile <- system.file("extdata", "SimVal/DAT-1-MI-PMX-2.csv", package = "PMXForest")
dfData   <- read.csv(dataFile)
```

# 1. `getCovStats()`

`getCovStats()` returns a named list, one element per covariate. The element type
depends on the number of unique values:

* **continuous** (more than `minLevels` unique values) — a two-element numeric
  vector of quantiles, by default the 5th and 95th percentiles;
* **binary** — the two levels, sorted alphanumerically;
* **multi-level categorical** — a nested list of one-hot vectors, first level as
  the reference.

```{r}
covStats <- getCovStats(
  data       = dfData,
  covariates = c("WT", "AGE", "SEX", "GENO"),
  idVar      = "ID"
)
covStats
```

## One record per subject

Statistics are computed on the first record per `idVar`, so subjects with dense
sampling do not pull the quantiles. Pass the identifier column that matches the
data:

```{r}
nrow(dfData)                                   # all records
nrow(distinct(dfData, ID))                     # subjects
```

## `minLevels`

`minLevels` (default 10) is the threshold between categorical and continuous. A
covariate coded with a small number of distinct numeric values — a dose level, an
ordinal score — is treated as categorical unless `minLevels` is lowered.

```{r}
## GENO has 4 levels: categorical by default
getCovStats(dfData, "GENO", idVar = "ID")

## Force GENO to be treated as continuous
getCovStats(dfData, "GENO", idVar = "ID", minLevels = 3)
```

## `probs` and `nsig`

`probs` sets the quantiles for continuous covariates; `nsig` sets the rounding.

```{r}
getCovStats(dfData, "CRCL", idVar = "ID", probs = c(0.1, 0.9), nsig = 4)
```

::: {.callout-note}
Binary levels are returned in alphanumeric order (`c(0, 1)`), not order of
appearance in the data. Forest plot row order is therefore reproducible across
data sets.
:::

# 2. `createInputForestData()`

`createInputForestData()` turns a covariate list into the row structure. It is
what you call directly when the covariate values are known literals rather than
derived from data.

```{r}
dfCovsManual <- createInputForestData(
  list(
    "WT"   = c(60, 100),
    "SEX"  = c(1, 2),
    "GENO" = list("GENO_2" = c(0, 1, 0), "GENO_3" = c(0, 0, 1), "GENO_4" = c(0, 0, 0))
  ),
  iMiss = -99
)
dfCovsManual
```

The `COVARIATEGROUPS` column records which covariate each row belongs to;
`getForestDFSCM()` uses it to group and space the rows on the y-axis, and drops it
before calling the parameter function.

A `getCovStats()` result can be passed straight in:

```{r}
createInputForestData(covStats, iMiss = -99)
```

# 3. `setupDfCovs()`

`setupDfCovs()` runs `getCovStats()` and `createInputForestData()` together and
adds control over how inactive cells are filled.

```{r}
covariates <- c("WT", "AGE", "CRCL", "SEX", "NCIL", "GENO")

dfCovs <- setupDfCovs(data = dfData, covariates = covariates, idVar = "ID")
dfCovs
```

## `additionalCovs`

Covariates listed in `additionalCovs` get their own rows, but on every *other*
row they sit at their data reference (mode for categorical, median or mean for
continuous) rather than at `-99`. This is for covariates that must hold a
plausible value whenever they are not the covariate under study — for example a
covariate carried through from a full random-effects (FREM) model.

```{r}
dfCovsAdd <- setupDfCovs(
  data           = dfData,
  covariates     = c("WT", "AGE", "SEX"),
  additionalCovs = "CRCL",
  idVar          = "ID"
)
dfCovsAdd
```

On the `WT`, `AGE`, and `SEX` rows, `CRCL` holds its median instead of `-99`.

## `useMissVal`

By default (`useMissVal = TRUE`) inactive *primary* covariates hold `-99`, and the
parameter function is responsible for treating that as "no effect". With
`useMissVal = FALSE`, inactive primary covariates are also set to their data
reference, so the parameter function never sees `-99`.

```{r}
dfCovsFilled <- setupDfCovs(
  data       = dfData,
  covariates = c("WT", "AGE", "SEX"),
  useMissVal = FALSE,
  idVar      = "ID"
)
dfCovsFilled
```

## `contRef`

`contRef` selects the reference statistic for continuous covariates, `"median"`
(default) or `"mean"`. It applies to `additionalCovs`, to primary covariates when
`useMissVal = FALSE`, and to `setupDfRefRow()`.

# 4. `setupDfRefRow()`

`setupDfRefRow()` builds the reference passed as `dfRefRow` to `getForestDFSCM()`.
It computes the reference state from the data — mode for categorical, median or
mean for continuous, on one record per subject — and maps it onto the geometry of
a `setupDfCovs()` result.

## `singleRef = TRUE`

A one-row data frame with every covariate at its reference value.

```{r}
refSingle <- setupDfRefRow(
  dfCovs     = dfCovs,
  data       = dfData,
  covariates = covariates,
  idVar      = "ID"
)
refSingle
```

## `singleRef = FALSE`

A data frame the same shape as `dfCovs`: active cells are replaced with the
reference value, inactive cells keep `-99`. Use this when the reference should
track the covariate that varies on each row — an age-specific reference for the
age rows, for instance.

```{r}
refMatrix <- setupDfRefRow(
  dfCovs     = dfCovs,
  data       = dfData,
  covariates = covariates,
  singleRef  = FALSE,
  idVar      = "ID"
)
refMatrix
```

::: {.callout-warning}
`setupDfRefRow()` keys the reference values by the column names `setupDfCovs()`
produces (`GENO_2`, `GENO_3`, …). Pass it a `dfCovs` built by `setupDfCovs()`. For
a hand-built `createInputForestData()` structure, construct `dfRefRow` directly as
a one-row data frame.
:::

# 5. Missing covariate values

`getCovStats()` and the setup functions exclude `missVal` before computing
statistics, so a covariate that is missing for some subjects is summarised from
the subjects who have it.

```{r}
dfMiss <- data.frame(
  ID  = 1:8,
  WT  = c(60.5, 70.2, 70.5, 80.8, 65.1, 90.0, 77.4, 68.9),
  SEX = c(0, 1, 1, 0, 1, 0, 1, 0),
  BMI = c(22.1, 24.4, -99, 27.3, 23.5, 28.9, -99, 25.1)
)

getCovStats(dfMiss, c("WT", "SEX", "BMI"), idVar = "ID", minLevels = 4)
```

The `BMI` quantiles are computed from the six non-missing values. The parameter
function still has to handle `-99` in the `BMI` column of `dfCovs`, because that
is what marks the rows where `BMI` is inactive.

# 6. Feeding the result forward

The data frame from `setupDfCovs()` (or `createInputForestData()`) is the `dfCovs`
argument of `getForestDFSCM()`; the data frame from `setupDfRefRow()` is its
`dfRefRow` argument. See the [Walkthrough](Part2-walkthrough.html) for the full
call. For empirical Forest plots, `getForestDFemp()` takes a list of covariate
expressions instead of `dfCovs` — the setup functions in this vignette are not
used there.
