---
title: "3. Deep Dive: Forest Plots for R-Coded Models"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Forest Plots for R-Coded Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(ggplot2)
  library(boot)
})

theme_set(theme_bw(base_size = 14))
theme_update(plot.title = element_text(hjust = 0.5))

set.seed(865765)
```

# Introduction

The parametric workflow in the [Quick-Start](Part1-quick-start.html) and
[Walkthrough](Part2-walkthrough.html) relies on `getForestDFSCM()`, which expects
NONMEM-style parameter vectors. When the model is fitted in R, the plot data can
be assembled directly and passed to `forestPlot()`.

This vignette builds a Forest plot from a logistic regression model fitted with
`glm()`, with the uncertainty taken from a non-parametric bootstrap using the
`boot` package. The response is survival in the `Titanic` data set.

```{r}
dfT <- as.data.frame(Titanic)

## Expand the frequency table to one row per passenger
dfData <- data.frame()
for (i in seq_len(nrow(dfT))) {
  if (dfT$Freq[i] != 0) {
    dfData <- rbind(
      dfData,
      data.frame(
        Class = as.character(dfT$Class[i]),
        Sex = as.character(dfT$Sex[i]),
        Age = as.character(dfT$Age[i]),
        Survived = as.character(dfT$Survived[i]),
        stringsAsFactors = TRUE
      )[rep(1, dfT$Freq[i]), ]
    )
  }
}
```

Three pieces are needed:

1. the covariate values to display;
2. the model, as R functions for fitting and prediction;
3. the uncertainty distribution of the predictions.

# 1. Fit and prediction functions

`glmfitpredfunc()` fits the logistic model on a bootstrap resample and predicts
the survival probability for a set of covariate combinations. Its signature
(`data`, `indices`, ...) is the form `boot::boot()` requires.

```{r}
## Fit a logistic regression
glmfitfunc <- function(fitdata, indices, strDV, strCovs, strInteraction = "") {
  dt <- fitdata[indices, ]
  glm(
    formula = paste0(strDV, " ~ ", paste(strCovs, collapse = "+"), strInteraction),
    family  = binomial(link = "logit"),
    data    = dt
  )
}

## Predict the probability of the modelled event
glmpredfunc <- function(glmfit, preddata) {
  predict(glmfit, newdata = preddata, type = "response")
}

## Fit and predict in one call
glmfitpredfunc <- function(fitdata, indices, strDV, strCovs,
                           strInteraction = "", preddata = NULL) {
  if (is.null(preddata)) preddata <- fitdata[indices, ]
  glmpredfunc(
    glmfitfunc(fitdata, indices, strDV, strCovs, strInteraction),
    preddata
  )
}
```

# 2. Covariate values

`createInputForestData()` builds the row structure. Each row of `dfCovs` is one
row of the Forest plot. For this categorical model, the inactive cells (`-99`) are
set to the reference level of each covariate, so every row is a complete covariate
combination that `predict()` can evaluate.

```{r}
dfCovs <- createInputForestData(
  list(
    "Class" = c("1st", "2nd", "3rd", "Crew"),
    "Sex"   = c("Male", "Female"),
    "Age"   = c("Child", "Adult")
  ),
  iMiss = -99
)
dfCovs$Class[dfCovs$Class == -99] <- "3rd"
dfCovs$Sex[dfCovs$Sex == -99] <- "Male"
dfCovs$Age[dfCovs$Age == -99] <- "Child"
dfCovs

covnames <- c(
  "First class", "2nd class", "3rd class", "Crew",
  "Male", "Female", "Child", "Adult"
)
```

The reference subject is row 3 — a boy travelling in 3rd class.

# 3. Assemble the plot data

`boot()` refits the model on 500 resamples and predicts the survival probability
for every row of `dfCovs` on each. The confidence interval is the 2.5th and
97.5th percentile of those predictions; the point estimate is the median.

```{r}
bootPreds <- boot(
  dfData,
  R = 500, statistic = glmfitpredfunc,
  strDV = "Survived", strCovs = c("Age", "Sex", "Class"),
  strInteraction = "", preddata = dfCovs
)

## Predictions from the fit on the full data
orgPred <- glmfitpredfunc(
  dfData, seq_len(nrow(dfData)),
  strDV = "Survived", strCovs = c("Class", "Age", "Sex"), strInteraction = ""
)
```

`forestPlot()` reads a fixed set of columns. Assembling them by hand means the
plot is not tied to any particular fitting engine.

```{r}
dfres <- dfCovs
dfres$COVNAME <- covnames
dfres$GROUP <- c(1, 1, 1, 1, 5, 5, 7, 7)
dfres$COVNUM <- seq_len(nrow(dfres))

## Reference from the bootstrap (row 3: boy, 3rd class) and from the observed data
dfres$REFFUNC <- 100 * median(bootPreds$t[, 3])
dfres$REFTRUE <- 100 * median(orgPred)

for (i in seq_len(nrow(dfres))) {
  q <- quantile(bootPreds$t[, i], probs = c(0.025, 0.5, 0.975), names = FALSE)
  dfres$Q1[i] <- 100 * q[1]
  dfres$POINT[i] <- 100 * q[2]
  dfres$Q2[i] <- 100 * q[3]
}

dfres$GROUPNAME <- dfres$COVARIATEGROUPS
dfres$PARAMETER <- "Chance of surviving Titanic %"
dfres$COVEFF <- FALSE
```

| Column | Meaning |
| --- | --- |
| `COVNAME`, `GROUPNAME` | row label and covariate-group label |
| `GROUP`, `COVNUM` | grouping index and row order |
| `POINT`, `Q1`, `Q2` | point estimate and confidence-interval limits |
| `REFFUNC` | reference value the relative scale divides by |
| `REFTRUE` | observed-data reference, drawn as the reference line |
| `PARAMETER` | facet label |

# 4. Draw the plot

The data is on an absolute scale (percent surviving), so `plotRelative = FALSE`.
The reference line is the observed median survival.

```{r ForestSurvival, fig.width=10, fig.asp=0.5, out.width="100%"}
forestPlot(
  dfres,
  plotRelative = FALSE,
  referenceInfo = NULL,
  xlb = "Reference subject is a boy in 3rd class",
  errbartabwidth = c(1, 1)
)
```

::: {.callout-note}
The same pattern works for any model that can be resampled: fit on each replicate,
predict for the covariate combinations, then fill `POINT`, `Q1`, `Q2`, and the
reference columns. For NONMEM parameter vectors, `getForestDFSCM()` does this
assembly for you — see the [Walkthrough](Part2-walkthrough.html).
:::
