---
title: "2. Walkthrough: End-to-End Forest Plots for NONMEM Models"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{2. Walkthrough: End-to-End Forest Plots for NONMEM Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.callout-note {
  background-color: #e7f1ff;
  border-left: 4px solid #0dcaf0;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
.callout-warning {
  background-color: #fff3cd;
  border-left: 4px solid #ffc107;
  color: #664d03;
  padding: 15px;
  margin-top: 20px;
  margin-bottom: 20px;
}
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 9
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(dplyr)
  library(ggplot2)
})

theme_set(theme_bw(base_size = 14))
theme_update(plot.title = element_text(hjust = 0.5))

set.seed(865765)
```

# Introduction

This walkthrough covers the full process of building Forest plots from a NONMEM
model and PsN output. It extends the [Quick-Start](Part1-quick-start.html) with:

* two ways to build the covariate structure, and how to label the rows;
* the parameter function in detail, including missing-value handling;
* the four sources of parameter uncertainty `getSamples()` accepts;
* an explicit reference subject with `setupDfRefRow()`;
* *empirical* Forest plots with `getForestDFemp()`;
* the formatting arguments of `forestPlot()`.

The example is a PK model (`run7`) built with the stepwise covariate model (SCM)
in PsN. Because SCM keeps only covariates that meet a significance criterion, some
rows of a parametric Forest plot from an SCM model show no effect. An empirical
Forest plot instead reads the covariate effect from the analysis data and can
therefore show covariates that are not in the model.

```{r}
modDir <- "SimVal"
dataFile <- system.file("extdata", paste0(modDir, "/DAT-1-MI-PMX-2.csv"), package = "PMXForest")
extFile <- system.file("extdata", paste0(modDir, "/run7.ext"), package = "PMXForest")
covFile <- system.file("extdata", paste0(modDir, "/run7.cov"), package = "PMXForest")
bsFile <- system.file("extdata", paste0(modDir, "/bs7.dir/raw_results_run7bs.csv"), package = "PMXForest")
sirFile <- system.file("extdata", paste0(modDir, "/sir7.dir/raw_results_run7.csv"), package = "PMXForest")

dfData <- read.csv(dataFile)
```

# 1. Build the covariate structure

Each row of `dfCovs` becomes one row of the Forest plot. The covariate effect on
that row is a function of the covariate values in that row. Cells that are not
active on a row hold the missing-value indicator (`-99` by default).

## 1.1 Explicit: `createInputForestData()`

`createInputForestData()` turns a named list of covariate values into the row
structure. A nested list defines a multi-level categorical covariate as a set of
one-hot columns.

The `run7` model uses genotype level 2 (2D6 EM) as its reference and estimates
effects for levels 1, 3, and 4. Its control stream names the indicator variables
`GENO1`, `GENO3`, `GENO4` (no separator), and the analysis data set carries
columns of the same names, so the one-hot columns are named to match and the
all-zero row is level 2.

```{r}
dfCovs <- createInputForestData(
  list(
    "WT" = c(65, 115),
    "AGE" = c(27, 62),
    "CRCL" = c(83, 150),
    "SEX" = c(1, 2),
    "NCIL" = c(0, 1),
    "FORM" = c(0, 1),
    "FOOD" = c(0, 1),
    "GENO" = list(
      "GENO1" = c(1, 0, 0, 0),
      "GENO3" = c(0, 0, 1, 0),
      "GENO4" = c(0, 0, 0, 1)
    )
  ),
  iMiss = -99
)
dfCovs
```

Row labels and covariate-group labels are supplied separately. Keep the label
vector length locked to the number of rows so a miscount fails immediately rather
than shifting every label below it.

```{r}
covnames <- c(
  "WT 65 kg", "WT 115 kg",
  "Age 27 y", "Age 62 y",
  "CRCL 83 mL/min", "CRCL 150 mL/min",
  "Male", "Female",
  "NCI = 0", "NCI > 0",
  "Oral tablet", "FDC",
  "Fasted", "Fed",
  "2D6 UM", "2D6 EM", "2D6 IM", "2D6 PM"
)
stopifnot(length(covnames) == nrow(dfCovs))

covariateGroupNames <- c(
  "Weight", "Age", "Creatinine\nclearance", "Sex",
  "NCI", "Formulation", "Food status", "2D6 genotype"
)
```

## 1.2 Streamlined: `setupDfCovs()`

`setupDfCovs()` derives the covariate values from the data instead of taking them
as literals. It summarises each covariate (`getCovStats()`) and reshapes the
result (`createInputForestData()`) in one call. Continuous covariates are
represented by their 5th and 95th percentiles; categorical covariates are
expanded to one row per level.

```{r}
dfCovsAuto <- setupDfCovs(
  data       = dfData,
  covariates = c("WT", "AGE", "CRCL", "SEX", "NCIL", "FORM", "FOOD", "GENO", "RACEL2"),
  catRef     = list(GENO = 2),
  sep        = "",
  idVar      = "ID"
)
dfCovsAuto
```

By default `setupDfCovs()` drops the lowest level of a categorical covariate as
the reference and names the one-hot columns `<covariate>_<level>`.
`catRef = list(GENO = 2)` makes 2D6 EM the reference, and `sep = ""` names the
columns `GENO1`, `GENO3`, `GENO4` — matching the `run7` control stream, the
analysis data set, and the explicit `dfCovs` from Section 1.1. The same parameter
function then serves both. `setupDfCovs()` does not produce row labels; supply
`cdfCovsNames` yourself, or leave it at its default to have each row labelled from
its covariate values. The `additionalCovs`, `useMissVal`, and `contRef` arguments
are covered in the
[Forest plot inputs deep dive](Part3-deep-dive-forest-plot-inputs.html).

The rest of this walkthrough uses the explicit `dfCovs` from Section 1.1.

# 2. Write the parameter function

The parameter function receives a vector of `thetas`, in the order the THETAs
appear in the NONMEM model, and a one-row data frame `df` of covariate values. It
rebuilds the relevant part of the model and returns a list of the quantities to
plot.

Two defensive checks apply to every covariate:

* `any(names(df) == "<cov>")` — the column may be absent entirely.
* `df$<cov> != -99` — the cell may hold the missing-value indicator, which
  `dfCovs` uses for covariates that are inactive on a row. This is needed even
  when the original NONMEM data has no missing covariate values.

Genotype is read from the one-hot columns `GENO1`, `GENO3`, `GENO4` — the same
names the NONMEM control stream uses. The parametric `dfCovs` carries them by
construction; in the empirical Forest plot (Section 6) they come from the raw
`GENO` column in the analysis data.

```{r}
paramFunction <- function(thetas, df, ...) {
  CLFOOD <- 1
  if (any(names(df) == "FOOD") && df$FOOD != -99 && df$FOOD == 0) CLFOOD <- 1 + thetas[11]

  FRELFORM <- 1
  if (any(names(df) == "FORM") && df$FORM != -99 && df$FORM == 0) FRELFORM <- 1 + thetas[12]

  FRELSEX <- 1
  if (any(names(df) == "SEX") && df$SEX != -99 && df$SEX == 2) FRELSEX <- 1 + thetas[14]

  FRELGENO4 <- 1
  if (any(names(df) == "GENO4") && df$GENO4 != -99 && df$GENO4 == 1) FRELGENO4 <- 1 + thetas[13]

  FREL <- thetas[1] * FRELSEX * FRELFORM * FRELGENO4

  if (any(names(df) == "WT") && df$WT != -99) {
    TVCL <- thetas[4] * (df$WT / 75)^thetas[2]
  } else {
    TVCL <- thetas[4]
  }
  if (any(names(df) == "GENO1") && df$GENO1 != -99 && df$GENO1 == 1) TVCL <- TVCL * (1 + thetas[8])
  if (any(names(df) == "GENO3") && df$GENO3 != -99 && df$GENO3 == 1) TVCL <- TVCL * (1 + thetas[9])
  if (any(names(df) == "GENO4") && df$GENO4 != -99 && df$GENO4 == 1) TVCL <- TVCL * (1 + thetas[10])
  CL <- CLFOOD * TVCL

  if (any(names(df) == "WT") && df$WT != -99) {
    V <- thetas[5] * (df$WT / 75)^thetas[3]
  } else {
    V <- thetas[5]
  }

  ## Secondary parameter: AUC after an 80 mg dose
  AUC <- 80 / (CL / FREL)

  list(CL, FREL, AUC, V)
}

functionListName <- c("CL", "Frel", "AUC", "V")
```

# 3. Sample the parameter uncertainty

`getSamples()` returns a data frame of parameter vectors in `.ext` column order.
The confidence interval on each Forest plot row is the quantile of the parameter
function evaluated over these vectors.

```{r}
## Covariance matrix: n draws from the multivariate normal implied by the .cov step
dfSamplesCOV <- getSamples(covFile, extFile = extFile, n = 200)

## Bootstrap: the non-zero-OFV resamples from a PsN bootstrap raw_results file
dfSamplesBS <- getSamples(bsFile, extFile = extFile)

## SIR: all resamples in a PsN SIR raw_results file
dfSamplesSIR <- getSamples(sirFile, extFile = extFile)

## Small bootstrap: build a covariance matrix from the bootstrap resamples,
## then draw n vectors from it
dfSamplesBSsmall <- getSamples(bsFile, extFile = extFile, n = 200)
```

Sampling from the covariance matrix or from a small bootstrap produces symmetric
intervals. A full bootstrap or SIR file lets the interval be asymmetric.

# 4. Define the reference subject

Without a reference, `getForestDFSCM()` uses the model with every covariate effect
switched off (all cells `-99`). To make the reference a typical subject instead,
pass `dfRefRow`.

`setupDfRefRow()` builds that row from the data, matched to the geometry of a
`setupDfCovs()` result: the mode for categorical covariates, the median (or mean)
for continuous covariates, computed on one record per subject.

```{r}
dfRefRow <- setupDfRefRow(
  dfCovs     = dfCovsAuto,
  data       = dfData,
  covariates = c("WT", "AGE", "CRCL", "SEX", "NCIL", "FORM", "FOOD", "GENO", "RACEL2"),
  catRef     = list(GENO = 2),
  sep        = "",
  idVar      = "ID"
)
dfRefRow
```

Pass the same `catRef` and `sep` here as to `setupDfCovs()` so the reference
row's one-hot columns line up.

::: {.callout-note}
The reference above is taken from the data - the median for a continuous
covariate, the most common level for a categorical one. If the parameter
function instead falls back to the model's own reference when a covariate is
inactive, as one built by `createParamFunction()` does, the two disagree and
every row where that covariate is inactive is displaced from 1. Setting
`contRef = "model"` and `catRef = "model"`, with `model` pointing at the control
stream, takes both from `$PK` and removes the mismatch. See the
[Forest plot inputs deep dive](Part3-deep-dive-forest-plot-inputs.html).
:::

For an explicit `dfCovs`, a one-row data frame works just as well:

```{r}
dfRefRowManual <- data.frame(
  WT = 75, AGE = 45, CRCL = 100, SEX = 1,
  NCIL = 0, FORM = 1, FOOD = 1,
  GENO1 = 0, GENO3 = 0, GENO4 = 0
)
```

::: {.callout-note}
The reference enters the relative scale as the denominator: each row is plotted as
its parameter value divided by the reference-subject value. A reference subject
that carries a large covariate effect shifts every other row. Choose it
deliberately, and state it in the plot caption.
:::

# 5. Compute the parametric Forest plot data

`getForestDFSCM()` evaluates `paramFunction` for every row of `dfCovs` and every
parameter vector, then appends the summary columns `forestPlot()` reads.
`noBaseThetas` is the number of THETAs in the model. `dfCovs` here already carries
the `GENO1`/`GENO3`/`GENO4` columns; a `dfCovs` with a raw `GENO` column instead
can be encoded on the fly with `oneHot = list(GENO = list(ref = 2))` and
`oneHotSep = ""`.

```{r}
dfresBS <- getForestDFSCM(
  dfCovs           = dfCovs,
  cdfCovsNames     = covnames,
  functionList     = list(paramFunction),
  functionListName = functionListName,
  noBaseThetas     = 14,
  dfParameters     = dfSamplesBS,
  dfRefRow         = dfRefRowManual
)
head(dfresBS[, c("COVNUM", "COVNAME", "GROUPNAME", "PARAMETER", "POINT", "Q1", "Q2")])
```

Switching the uncertainty source is a one-argument change:

```{r}
dfresCOV <- getForestDFSCM(
  dfCovs           = dfCovs,
  cdfCovsNames     = covnames,
  functionList     = list(paramFunction),
  functionListName = functionListName,
  noBaseThetas     = 14,
  dfParameters     = dfSamplesCOV,
  dfRefRow         = dfRefRowManual
)
```

```{r}
forestPlot(dfresBS, parameters = "CL", groupNameLabels = covariateGroupNames)
```

# 6. Empirical Forest plots

An empirical Forest plot reads each covariate effect from the analysis data. The
covariate structure is a list of expressions that subset the data, and the
parameter function is the same one used above.

The parameter function needs the `GENO1`/`GENO3`/`GENO4` columns. This analysis
data set already contains them, but passing `oneHot = list(GENO = list(ref = 2))`
with `oneHotSep = ""` builds them from the raw `GENO` column when they are not
present, and is a harmless no-op when they are.

```{r}
lsExpr <- list(
  "NCI"  = expression(NCIL == 0),
  "NCI"  = expression(NCIL == 1),
  "FORM" = expression(FORM == 0),
  "FORM" = expression(FORM == 1),
  "FOOD" = expression(FOOD == 0),
  "FOOD" = expression(FOOD == 1),
  "GENO" = expression(GENO == 1),
  "GENO" = expression(GENO == 2),
  "GENO" = expression(GENO == 3),
  "GENO" = expression(GENO == 4),
  "WT"   = expression(WT < 70),
  "WT"   = expression(WT > 104),
  "SEX"  = expression(SEX == 1),
  "SEX"  = expression(SEX == 2)
)

covnamesEmp <- c(
  "NCI = 0", "NCI > 0", "Oral tablet", "FDC", "Fasted", "Fed",
  "2D6 UM", "2D6 EM", "2D6 IM", "2D6 PM",
  "WT < 70 kg", "WT > 104 kg", "Male", "Female"
)
stopifnot(length(covnamesEmp) == length(lsExpr))
```

`getForestDFemp()` needs the analysis data. Empirical plots are more expensive to
compute than parametric ones, so a subset keeps the example quick; `ncores > 1`
parallelises the calculation and `cstrPackages` names the packages the workers
need.

```{r}
dfDataEmp <- read.csv(dataFile) %>%
  distinct(ID, .keep_all = TRUE) %>%
  slice(1:300)

dfresEmp <- getForestDFemp(
  dfData             = dfDataEmp,
  covExpressionsList = lsExpr,
  cdfCovsNames       = covnamesEmp,
  functionList       = list(paramFunction),
  functionListName   = functionListName,
  noBaseThetas       = 14,
  dfParameters       = dfSamplesCOV,
  oneHot             = list(GENO = list(ref = 2)),
  oneHotSep          = "",
  ncores             = 1,
  cstrPackages       = "dplyr"
)
```

```{r}
forestPlot(dfresEmp, parameters = "CL", noVar = FALSE)
```

`noVar = FALSE` keeps the parameter-uncertainty interval. `noVar = TRUE` (the
default for empirical data) drops it and shows only the covariate effect.

# 7. Customise the plot

`forestPlot()` returns a `ggplot` object (or a list of them). Common arguments:

| Argument | Effect |
| --- | --- |
| `parameters` | which of `functionListName` to draw, and in what order |
| `groupNameLabels` | replace the covariate-group strip labels |
| `plotRelative` | `TRUE` (default) plots relative to the reference; `FALSE` plots the absolute value |
| `noVar` | drop the parameter-uncertainty interval |
| `xlim` | x-axis limits |
| `statisticsLabel` | prefix for the statistics-panel strip labels |
| `return` | `"plot"` (default) or `"plotList"` for the panel list |

```{r}
forestPlot(
  dfresBS,
  parameters      = c("CL", "Frel"),
  groupNameLabels = covariateGroupNames,
  statisticsLabel = "Estimate:",
  xlim            = c(0.5, 1.6)
)
```

# Where to next

* [Preparing Forest plot inputs](Part3-deep-dive-forest-plot-inputs.html) —
  all four inputs in full: the covariate structure, generating the parameter
  function from the control stream, the reference row, and keeping the last two
  consistent with each other.
* [R-coded models](Part3-deep-dive-r-coded-models.html) — Forest plots for models
  fitted outside NONMEM, with the plot data assembled by hand.
* [Time-to-event models](Part3-deep-dive-tte-models.html) — hazard ratios, event
  probabilities, and distribution reparameterisation.
