---
title: "3. Deep Dive: Secondary Parameters with mrgsolve"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: default
vignette: >
  %\VignetteIndexEntry{3. Deep Dive: Secondary Parameters with mrgsolve}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  # Simulation and bootstrap chunks are shown but not run when the vignette is
  # built: mrgsolve needs a C++ toolchain and the Forest data set takes minutes.
  # Flip this to TRUE (with mrgsolve installed) to execute end to end.
  eval     = eval
)
```

```{r setup}
suppressPackageStartupMessages({
  library(PMXForest)
  library(ggplot2)
})
theme_set(theme_bw(base_size = 14))
set.seed(865765)
```

# Introduction

A Forest plot usually shows model parameters -- `CL`, `V`, a bioavailability
term. Often the quantity of clinical interest is a **secondary** parameter
derived from those: steady-state `Cmax`, an exposure `AUC` over a dosing
interval, a time above some threshold. These are not in `$PK`; they come from
solving the model for a particular dosing design.

`createParamFunction()` translates `$PK` into an R parameter function
automatically. Its `secondary` argument lets you attach the derived quantities
in the same call, so the whole parameter function -- structural part and
secondary part -- is generated and stays in one object that
`getForestDFSCM()` can consume directly.

Each `secondary` entry is a named string that is **either**

* a line of R code -- `secondary = list(AUC = "df$DOSE / CL")` -- or
* the path to an `.R` file of arbitrary code, including an `mrgsolve` model and
  simulation -- `secondary = list(CMAX = "cmax.R")`.

A file's text is *inlined* into the generated source, so the result stays a
self-contained artifact: it does not depend on the file still being on disk when
the parameter function later runs on a compute node.

This vignette builds a Forest plot for **steady-state Cmax after q24h oral
dosing**, with the Cmax computed by an `mrgsolve` simulation.

# 1. The structural parameter function

We start from the bundled `run7` control stream and ask for the three structural
parameters we care about.

```{r}
modFile <- system.file("extdata", "SimVal/run7.mod", package = "PMXForest")
extFile <- system.file("extdata", "SimVal/run7.ext", package = "PMXForest")

base <- suppressWarnings(
  createParamFunction(modFile, parameters = c("CL", "V", "MAT"), quiet = TRUE)
)
base$functionListName
```

The generated function returns typical values of `CL`, `V` and `MAT`. `run7`'s
`$PK` also computes `FREL` (relative bioavailability) and `KA`, and the
generator keeps those intermediate assignments -- which means a secondary block
spliced in afterwards can use `FREL` and `KA` by name.

# 2. The mrgsolve secondary

The file below is shipped with the package as
`system.file("secondary-cmax-mrgsolve.R", ...)` for the vignette. It is ordinary
R: it compiles a one-compartment oral model with `mrgsolve`, simulates seven
q24h doses, and returns the maximum concentration in the final interval. The
**value of its last expression** becomes `CMAX`.

```{r, echo = FALSE, comment = ""}
cmaxFile <- system.file("secondary-cmax-mrgsolve.R", package = "PMXForest")
cat(readLines(cmaxFile), sep = "\n")
```

Points worth noting:

* **Scope.** Inside the generated function the block sees `thetas`, `df`, `...`
  and every structural parameter (`CL`, `V`, `MAT`, `FREL`, `KA`, ...) by name.
  Covariate columns are reached as `df$NAME` -- there is no bare `WT`.
* **Compilation cost.** `mrgsolve::mcode_cache()` keeps the compiled model
  between calls. The parameter function is evaluated once per covariate row per
  bootstrap sample -- hundreds to thousands of times -- so compiling on every
  call would dominate the run. Cache the model, or build it once outside and
  pass it in through `...`.
* **Return a scalar.** `getForestDFSCM()` expects each returned element to be a
  single number. The last line reduces the simulation to `max(...)`.

# 3. Generate the combined parameter function

```{r}
cmaxFile <- system.file("secondary-cmax-mrgsolve.R", package = "PMXForest")

out <- suppressWarnings(createParamFunction(
  modFile,
  parameters = c("CL", "V", "MAT"),
  secondary  = list(CMAX = cmaxFile),
  quiet      = TRUE
))

out$functionListName   # CL, V, MAT, then the secondary
out$primaryNames
out$secondaryNames
```

The tail of the generated source now carries the inlined simulation and a
return list extended with `CMAX`. The file's text is spliced in **verbatim**
(re-indenting it would corrupt the multi-line `mrgsolve` model string):

```r
  ## ---- Secondary parameters -----------------------------------------------
  ## CMAX  (inlined from secondary-cmax-mrgsolve.R)
  CMAX <- local({
.code <- "
$PARAM CL = 1, V = 10, KA = 1
...
"
.mod <- mrgsolve::mcode_cache("pmxforest_secondary_1cmt_oral", .code)
# ...
.lastInterval <- subset(.sim, time >= (.ndose - 1) * .tau)
max(.lastInterval$CP)
  })

  list(
    CL = CL,
    V = V,
    MAT = MAT,
    CMAX = CMAX
  )
```

## Sanity-check at the final estimates

Read the function in and call it at the `run7` final estimates for a reference
subject. `CMAX` should be a plausible steady-state peak.

```{r}
paramFunction <- eval(parse(text = out$code))

thetas <- as.numeric(
  getExt(extFile)[getExt(extFile)$ITERATION == -1000000000, 2:15]
)

paramFunction(thetas = thetas, df = data.frame(WT = 75, FOOD = 1))
#> $CL
#> [1] 11.9074
#>
#> $V
#> [1] 191.231
#>
#> $MAT
#> [1] 6.46559
#>
#> $CMAX
#> [1] 0.3699538
```

# 4. Feed it to `getForestDFSCM()`

Nothing special is needed: `functionListName` already carries `CMAX`, so the
Forest data frame gets a `CMAX` parameter for free. `mrgsolve` must be loaded on
the worker processes, hence `cstrPackages`.

```{r}
dfData <- read.csv(
  system.file("extdata", "SimVal/DAT-1-MI-PMX-2.csv", package = "PMXForest")
)
dfCovs <- setupDfCovs(
  dfData, covariates = c("WT", "FOOD", "FORM"), idVar = "ID"
)

dfSamples <- getSamples(
  system.file("extdata", "SimVal/run7.cov", package = "PMXForest"),
  extFile, n = 200
)

dfres <- getForestDFSCM(
  dfCovs           = dfCovs,
  functionList     = list(paramFunction),
  functionListName = out$functionListName,   # CL, V, MAT, CMAX
  noBaseThetas     = out$noBaseThetas,
  dfParameters     = dfSamples,
  cstrPackages     = "mrgsolve",
  ncores           = 1
)

unique(dfres$PARAMETER)
#> [1] "CL"   "V"    "MAT"  "CMAX"
```

# 5. The Forest plot

```{r, fig.width = 9, fig.asp = 0.7, out.width = "100%"}
forestPlot(
  dfres[dfres$PARAMETER == "CMAX", ],
  plotRelative = TRUE,
  keepYlabs    = TRUE,
  rightStrip   = FALSE
) +
  ggtitle("Steady-state Cmax vs. reference")
```

# Notes

* **Verification.** `verifyParamFunction()` checks the generated function
  against a NONMEM `$TABLE`. It checks only the `$PK` parameters by default --
  `CMAX` is not a table column and not reducible to a typical value -- so a
  function with secondaries still verifies cleanly. Pass
  `parameters = "CMAX"` to force a check if you do have it tabled.
* **AUC as well.** Add a second entry, e.g.
  `secondary = list(CMAX = "cmax.R", AUC = "auc.R")`, where `auc.R` integrates
  the concentration (an `$CMT AUC` compartment in the mrgsolve model, or
  `pracma::trapz()` on the simulated profile). Entries are evaluated in order,
  so `auc.R` could also reuse a value assigned by `cmax.R`.
* **Snippets vs. files.** A string is treated as a file when it exists on disk.
  A string that looks like a path (`.R` / `.r`) but does not exist is an error,
  not a silently mistyped snippet.
